---
title: "Differential expression LGR5+ Colon vs SI"
author: "Jason T. Serviss"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r, message = FALSE}
#PACKAGES
packages <- c("sp.scRNAseq", "sp.scRNAseqData", "ggthemes", "tidyverse", "seqTools")
purrr::walk(packages, library, character.only = TRUE)
rm(packages)
```

### Goal

Identify DE genes between colon and SI stem cells via single cell RNA-seq data from mouse.
This was perfored using 2 different strategies: a) pairwise comparison of all cell types
as defined by unsupervised classification and b) comparison of the top 50 Lgr5+ colon vs
small intestine cells from the stem cell population (as defined above).

### Run unsupervised classification

First we run unsupervised classificaiton using t-SNE and the Mclust software.

```{r}
#setup spCounts
s <- str_detect(colnames(countsMgfp), "^s")
countsMgfpMeta <- countsMgfpMeta %>%
mutate(mouseName = if_else(
  str_detect(plate, "NJA004"),
  "C57black/6J", "C57black/6J/Lgr5gfp"
  ))
 m <- colnames(countsMgfp) %in% filter(countsMgfpMeta, mouseName == "C57black/6J")$sample

cObjSng <- spCounts(countsMgfp[, s & m], cbind(countsMgfpERCC[, s & m]))
cObjMul <- spCounts(countsMgfp[, !s & m], cbind(countsMgfpERCC[, !s & m]))

#spUnsupervised
load("~/Github/sp.scRNAseqTesting/inst/differentialExpLgr5ColonSI/uObj.rda")
```

Plot unsupervised classification results.
```{r, fig.align="center", fig.width=10, fig.height=8}
plotUnsupervisedClass(uObj, cObjSng)
```

Show marker expression specific for the cell types in the tissue.
```{r, fig.align="center", fig.width=10, fig.height=8}
plotUnsupervisedMarkers(
  uObj, cObjSng,
  c("Lgr5", "Muc2", "Ptprc", "Chga", "Alpi", "Lyz1", "Dclk1", "Slc40a1"),
  pal = RColorBrewer::brewer.pal(8, "Set1")
)
```

Show the tissues.
```{r, fig.align="center", fig.width=10, fig.height=8}
plotUnsupervisedClass(uObj, cObjSng) %>%
plotData() %>%
inner_join(countsMgfpMeta, by = c("Sample" = "sample")) %>%
ggplot() +
geom_point(aes(`t-SNE dim 1`, `t-SNE dim 2`, colour = tissue)) +
theme_few() +
theme(legend.position = "top") +
guides(colour = guide_legend(title = "Tissue"))
```

Rename the classes
```{r, fig.align="center", fig.width=10, fig.height=8}
classifications <- tibble(
  sample = rownames(getData(uObj, "tsne")),
  oldClass = getData(uObj, "classification")
) %>%
mutate(newClass = case_when(
  oldClass == "C1" ~ "C.Stem",
  oldClass == "F1" ~ "C.Colonocyte",
  oldClass == "B1" ~ "C.Goblet",
  oldClass == "G1" ~ "Blood.Tufft.Endocrine",
  oldClass == "D1" ~ "SI.Stem",
  oldClass == "A1" ~ "SI.Enterocyte",
  oldClass == "E1" ~ "SI.Goblet",
  TRUE ~ "error"
))

classification(uObj) <- classifications$newClass[match(rownames(getData(uObj, "tsne")), classifications$sample)]
groupMeans(uObj) <- averageGroupExpression(cObjSng, getData(uObj, "classification"), FALSE)
tsneMeans(uObj) <- tsneGroupMeans(getData(uObj, "tsne"), getData(uObj, "classification"))
plotUnsupervisedClass(uObj, cObjSng)
```

Show Lgr5 expression.

```{r, fig.align="center", fig.width=10, fig.height=8}
plotUnsupervisedMarkers(uObj, cObjSng, "Lgr5")
```

Expression is log2(cpm) and then scaled to the interval [0, 1] with: (exp - min(exp)) / (max(exp) - min(exp)).

Show Lgr5 expression per class and tissue.
```{r, fig.align="center", fig.width=10, fig.height=8}
plotUnsupervisedMarkers(uObj, cObjSng, "Lgr5") %>%
plotData() %>%
inner_join(countsMgfpMeta, by = c("Sample" = "sample")) %>%
ggplot() +
geom_histogram(
  aes(Lgr5, fill = tissue, y = ..count.., group = tissue), binwidth = 0.05, position = position_dodge(),
  alpha = 0.7
) +
facet_wrap(~Classification, scales = "free_y") +
theme_bw() +
theme(legend.position = "top", plot.caption = element_text(hjust = 0)) +
guides(fill = guide_legend(title = "Tissue")) +
labs(
  x = "Lgr5 expression: log2(cpm) scaled to the interval [0, 1] with: (exp - min(exp)) / (max(exp) - min(exp))."
)
```

### Differential expression strategy 1

Pairwise KStest using all classes.

```{r, eval = FALSE}
ks.res <- KStest(getData(cObjSng, "counts.cpm"), getData(uObj, "classification"), cores = 8)
p.ks.res <- processKStest(ks.res, getData(uObj, "classification"), 0.05) %>%
filter(sigBool) %>%
group_by(gene) %>%
filter(n() == 1) %>%
ungroup() %>%
mutate(p.prod = map_dbl(p.values, prod)) %>%
mutate(meanCpm = map2_dbl(id, gene, function(i, g) {
  mean(getData(cObjSng, "counts.cpm")[g, getData(uObj, "classification") == i])
})) %>%
mutate(log2.fc = map2_dbl(id, gene, function(i, g) {
  log2(
    mean(getData(cObjSng, "counts.cpm")[g, getData(uObj, "classification") == i]) /
    mean(getData(cObjSng, "counts.cpm")[g, getData(uObj, "classification") != i])
  )
})) %>%
arrange(desc(log2.fc), p.prod, desc(statSum), desc(meanCpm))

save(p.ks.res, file = './differentialExpressionStrategy1.rda', compress = "bzip2")
openxlsx::write.xlsx(p.ks.res, file = "./differentialExpressionStrategy1.xlsx")
```

```{r, echo = FALSE}
load('~/Github/sp.scRNAseqTesting/inst/differentialExpLgr5ColonSI/differentialExpressionStrategy1.rda')
load('~/Github/sp.scRNAseqTesting/inst/differentialExpLgr5ColonSI/differentialExpressionStrategy.ks.res1.rda')
```


Number of significant genes detected at alpha 0.05?

```{r}
p.ks.res %>% count(id)
```

Number of significant genes detected at alpha 0.01?

```{r}
processKStest(ks.res, getData(uObj, "classification"), 0.01) %>%
filter(sigBool) %>%
group_by(gene) %>%
filter(n() == 1) %>%
ungroup() %>%
mutate(p.prod = map_dbl(p.values, prod)) %>%
count(id)
```

Number of significant genes detected at alpha 0.001?

```{r}
processKStest(ks.res, getData(uObj, "classification"), 0.001) %>%
filter(sigBool) %>%
group_by(gene) %>%
filter(n() == 1) %>%
ungroup() %>%
mutate(p.prod = map_dbl(p.values, prod)) %>%
count(id)
```

Number of significant genes detected at alpha 0.05 and fold change > 2?

```{r}
p.ks.res %>%
filter(log2.fc > 1 | log2.fc < -1) %>%
count(id)
```

Plot some candidates in the context of all cells.

```{r}
#dotplot function
dotplot <- function(data) {
  data %>%
  rename(Gene = gene) %>%
  ggplot() +
  geom_dotplot(
    data = . %>% filter(expression != 0),
    aes(Classification, expression, colour = Gene, fill = Gene, group = interaction(Classification, Gene)),
    binaxis="y", stackdir="center", binwidth = 0.01, position = position_dodge()
  ) +
  geom_text(
    data = . %>%
      group_by(Classification, Gene) %>%
      summarize(do = paste0(round(length(which(expression == 0)) / n(), digits = 2) * 100, "%")),
    aes(Classification, 0.05, label = do), size = 2
  ) +
  geom_point(
    data = . %>%
      group_by(Gene, Classification) %>%
      summarize(median = median(expression)) %>%
      filter(median != 0),
    aes(Classification, median), shape = 95, size = 3
  ) +
  ylim(0, 1) +
  theme_few() +
  theme(axis.text.x = element_text(angle = 90), legend.position = "top") +
  facet_wrap(~Gene) +
  labs(
    x = "Tissue",
    y = "log2(expression)[0, 1]"
  ) +
  guides(fill = FALSE, colour = FALSE)
}
```

SI stem genes
```{r, fig.align="center", fig.width=10, fig.height=8}
#plot SI candidates
p.ks.res %>%
filter(id == "SI.Stem") %>%
arrange(desc(log2.fc), p.prod, desc(statSum), desc(meanCpm)) %>%
slice(1:30) %>%
pull(gene) %>%
plotUnsupervisedMarkers(uObj, cObjSng, .) %>%
plotData() %>%
gather(gene, expression, -(Sample:Colour)) %>%
dotplot()
```

Colon stem genes
```{r, fig.align="center", fig.width=10, fig.height=8}
p.ks.res %>%
filter(id == "C.Stem") %>%
arrange(desc(log2.fc), p.prod, desc(statSum), desc(meanCpm)) %>%
slice(1:30) %>%
pull(gene) %>%
plotUnsupervisedMarkers(uObj, cObjSng, .) %>%
plotData() %>%
gather(gene, expression, -(Sample:Colour)) %>%
dotplot()
```

Results were saved to the file "differentialExpressionStrategy1" with the following columns:

Id: the cell type that the corresponding gene has been found to be differentially expressed in.
Gene: the name of the gene
sigBool: a logical indicating if all pairwise tests were significant (alpha = 0.001)
statSum: the sum of test statistics from the KS tests
p.values: the individual p values from each pairwise test that was performed for the corresponding gene
Statistics: the individual test statistics from each pairwise test that was performed for the corresponding gene
p.prod: the product of the p.values from all pairwise tests
meanCpm: the mean counts per million expression values for the gene in the cell type listed in the “id” column
Log2.fc: the log 2 fold change comparing the cell type listed in the “id” column vs all other cell types as a group.

### Differential expression strategy 2

KStest using only the top 50 cells in each of the C.Stem and S.Stem classes with
highest Lgr5 and comparing only these classes.

```{r, eval = FALSE}
idx <- which(getData(uObj, "classification") %in% c("C.Stem", "SI.Stem"))
counts <- getData(cObjSng, "counts.cpm")[, idx]
classes <- getData(uObj, "classification")[idx]

#find top 50 cells in each class with highest Lgr5
top50 <- tibble(Lgr5 = counts["Lgr5", ], class = classes, sample = colnames(counts)) %>%
group_by(class) %>%
top_n(50, Lgr5) %>%
ungroup()

ks.res2 <- KStest(counts[, match(top50$sample, colnames(counts))], top50$class, cores = 8)
p.ks.res2 <- ks.res2 %>%
filter(p.value < 0.001) %>%
separate(combination, c("Celltype1", "Celltype2"), sep = "-")  %>%
mutate(
  meanCpmCelltype1 = map2_dbl(Celltype1, gene, function(i, g) {
    mean(getData(cObjSng, "counts.cpm")[g, getData(uObj, "classification") == i])
  }),
  meanCpmCelltype2 = map2_dbl(Celltype2, gene, function(i, g) {
    mean(getData(cObjSng, "counts.cpm")[g, getData(uObj, "classification") == i])
  })
) %>%
mutate(log2.fc = log2(meanCpmCelltype1 / meanCpmCelltype2)) %>%
arrange(desc(log2.fc), p.value)

save(p.ks.res2, file = '~/Github/sp.scRNAseq/inst/differentialExpLgr5ColonSI/differentialExpressionStrategy2.rda', compress = "bzip2")
openxlsx::write.xlsx(p.ks.res2, file = "~/Github/sp.scRNAseq/inst/differentialExpLgr5ColonSI/differentialExpressionStrategy2.xlsx")
```

```{r, echo = FALSE}
load('~/Github/sp.scRNAseqTesting/inst/differentialExpLgr5ColonSI/differentialExpressionStrategy2.rda')
load('~/Github/sp.scRNAseqTesting/inst/differentialExpLgr5ColonSI/differentialExpressionStrategy.ks.res2.rda')
```

Number of significant genes detected at alpha 0.05?

```{r}
ks.res2 %>%
filter(p.value < 0.05) %>%
count(combination)
```

Number of significant genes detected at alpha 0.01?

```{r}
ks.res2 %>%
filter(p.value < 0.01) %>%
count(combination)
```

Number of significant genes detected at alpha 0.001?

```{r}
ks.res2 %>%
filter(p.value < 0.001) %>%
count(combination)
```

Number of significant genes detected at alpha 0.05 and fold change > 2?

```{r}
p.ks.res2 %>%
filter(log2.fc > 1 | log2.fc < -1) %>%
unite(combination, Celltype1, Celltype2) %>%
count(combination)
```

Plot some candidates in the context of all cells.

SI stem genes
```{r, fig.align="center", fig.width=10, fig.height=8}
p.ks.res2 %>%
filter(log2.fc < 1) %>%
arrange(log2.fc, p.value) %>%
slice(1:30) %>%
pull(gene) %>%
plotUnsupervisedMarkers(uObj, cObjSng, .) %>%
plotData() %>%
gather(gene, expression, -(Sample:Colour)) %>%
dotplot()
```

Colon stem genes
```{r, fig.align="center", fig.width=10, fig.height=8}
p.ks.res2 %>%
filter(log2.fc > 1) %>%
arrange(desc(log2.fc), p.value) %>%
slice(1:30) %>%
pull(gene) %>%
plotUnsupervisedMarkers(uObj, cObjSng, .) %>%
plotData() %>%
gather(gene, expression, -(Sample:Colour)) %>%
dotplot()
```

Results	were saved to the file "differentialExpressionStrategy2" with the following columns:

Cell type 1: 1st cell type in comparison
Cell type 2: 2nd cell type in comparison
Gene: the name of the gene
Stat: the test statistic from the KS test that was performed for the corresponding gene
p.value: the p values from the KS test that was performed for the corresponding gene
meanCpmCellType1: mean counts per million for cell type 1
meanCpmCellType2: mean counts per million for cell type	2
Log2.fc: the log 2 fold change comparing the two cell types.

````{r}
sessionInfo()
```