---
title: "Connection Score"
author: "Jason T. Serviss"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r, message = FALSE}
packages <- c(
  "sp.scRNAseq", "sp.scRNAseqData", "printr",
  "ggthemes", "tidyverse"
)
purrr::walk(packages, library, character.only = TRUE)
rm(packages)
```


```{r}
#DATA
load('~/Github/sp.scRNAseqTesting/inst/testingPoissonSorted/data/uObj.rda')
load('~/Github/sp.scRNAseqTesting/inst/testingPoissonSorted/data/sObj.rda')

#FUNCTIONS
calculateScore <- function(sObj, samples = NULL) {
  swarm <- getData(sObj, "spSwarm")
  
  if(is.null(samples)) {
    samples <- rownames(swarm)
  }
  
  swarm <- as.data.frame(
    t(swarm[samples, ]), 
    stringsAsFactors = FALSE
  )
  costs <- getData(sObj, "costs")
  combos <- as.data.frame(combn(rownames(swarm), 2), stringsAsFactors = FALSE)
  prod <- map(combos, function(x) {
    swarm[x[1], ] * swarm[x[2], ]
  })
  byCost <- map(prod, function(x) x / costs)
  scores <- map_dbl(byCost, sum)
  as.data.frame(t(combos), stringsAsFactors = FALSE) %>%
    as_tibble() %>%
    rename(from = V1, to = V2) %>%
    add_column(score = scores)
}
```

The goal here was to develop a connection score that indicates the strength of 
the evidence for a connection in the data. This can, hopefully, subsequently be 
used in permutation tests to give p-value corresponding to the individual 
detected connections. Due to the fact that deconvolution of permuted data gives 
rise to high fractions, the score should be regulated by the cost.

To test the scoring function, we utilized the dataset with sorted multiplets. By
including an increasing amount of one connection type (multiplets that only 
have a HCT116-HOS connection) we can evaluate the behavior of the scoring 
funciton with a known number of connections.

```{r, fig.align="center", fig.width=10, fig.height=8}
#increase multiplets that ONLY have a HCT116-HOS connection
connectionSamples <- filter(countsSortedMeta2, cellTypes == "HCT116-HOS")$sample
allSamples <- rownames(getData(sObj, "spSwarm"))
otherSamples <- allSamples[!allSamples %in% connectionSamples]

include <- map(2:length(connectionSamples), function(x) {
  c(otherSamples, connectionSamples[1:x])
})
names(include) <- 2:length(connectionSamples)

map_dfr(seq_along(include), function(x) {
  calculateScore(sObj, samples = include[[x]])
}, .id = "realConnections") %>%
  unite(connection, from, to, sep = "-") %>%
  mutate(realConnections = as.numeric(realConnections)) %>%
  ggplot() +
  geom_point(aes(realConnections, score, colour = connection), size = 3) +
  scale_x_continuous(breaks = seq(1, 10, 1)) +
  theme_bw() +
  ggthemes::scale_colour_ptol() +
  labs(x = "Number of multiplets with only HCT116-HOS connections", y = "Score (Sum of fractions)") +
  guides(colour = guide_legend(title = "Connection", title.position = "top", title.hjust = 0.5)) +
  theme(legend.position = "top")
```

The results indicate that adding an increasing amount of HCT116-HOS connections
causes an increase in the HCT116-HOS connection score without effecting the 
scores of the other connections.

Several scoring functions were tested with the results below showing the 
results of one that exhibited the desired behavior. The scoring 
function works in the following fashion:

1. The product of the 2 fractions corresponding to the connection being 
evaluated is calculated for each multiplet. The product was used so that 
connections with fractions of the type 0, 0.5 will give a product/score of 0.

2. The products for each connection and multiplet are divided by the cost for
each multiplet.

3. The sum of these values is calculated for each connection.

So for each connection the following is calculated:

$$Score_{connection} = \sum_{i=1}^{n} \frac{(F_{ct1}\ from\ M_i * F_{ct2}\ from\ M_i)}{Cost\ from\ M_i}$$

Where: _F_ = fraction; _ct_ = cell type; _M_ = multiplet

```{r}
s <- seq(0, 1, 0.1)
c <- expand.grid(s, s)
combos <- c[rowSums(c) == 1, ]
combos
matrixStats::rowProds(as.matrix(combos))
```

