---
title: "Mouse analysis Enge Simple 5000 synthetic multiplets"
author: "Jason T. Serviss"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r, message = FALSE}
packages <- c("CIMseq", "CIMseq.data", "tidyverse", "scSeqTools", "printr")
purrr::walk(packages, library, character.only = TRUE)
rm(packages)

##DATA
load('../../MGA.analysis_engeSimple/data/CIMseqData.rda')
load('../data/sObj.rda')
```

```{r, eval = TRUE}
#rename classes
renameClasses <- function(class) {
  case_when(
    class == "0" ~ "C.Stem.proximal",
    class == "1" ~ "C.Stem.distal",
    class == "2" ~ "SI.Stem",
    class == "3" ~ "C.Goblet.distal",
    class == "4" ~ "SI.Enterocytes",
    class == "5" ~ "SI.Goblet",
    class == "6" ~ "C.Goblet.proximal",
    class == "7" ~ "Enteroendocrine",
    class == "8" ~ "Tufft",
    class == "9" ~ "SI.Paneth",
    class == "10" ~ "Blood",
    TRUE ~ "error"
  )
}

cOrder <- c(
  "C.Stem.proximal", "C.Goblet.proximal", "C.Stem.distal", "C.Goblet.distal",
  "SI.Goblet", "SI.Paneth", "SI.Stem", "SI.Enterocytes",
  "Enteroendocrine", "Tufft", "Blood"
)

getData(cObjSng, "classification") <- renameClasses(getData(cObjSng, "classification"))
features <- getData(cObjMul, "features")
names(features) <- renameClasses(names(features))
getData(cObjMul, "features") <- features
fractions <- getData(sObj, "fractions")
colnames(fractions) <- renameClasses(colnames(fractions))
sObj@fractions <- fractions
```

### Classification

```{r, fig.align="center", fig.width=10, fig.height=8, eval = TRUE}
plotUnsupervisedClass(cObjSng, cObjMul)
plotUnsupervisedMarkers(
  cObjSng, cObjMul,
  c("Lgr5", "Ptprc", "Chga", "Dclk1", "Alpi", "Slc26a3", "Atoh1", "Lyz1"),
  pal = RColorBrewer::brewer.pal(8, "Set1")
)
plotUnsupervisedMarkers(
  cObjSng, cObjMul, "Hoxb13",
  pal = RColorBrewer::brewer.pal(8, "Set1")
)
plotUnsupervisedMarkers(
  cObjSng, cObjMul, "Mki67",
  pal = RColorBrewer::brewer.pal(8, "Set1")
)
```

```{r, fig.align="center", fig.width=10, fig.height=8, eval = TRUE}
classHeatmap(
  data = data.frame(
    gene = rownames(getData(cObjMul, "counts"))[getData(cObjMul, "features")],
    class = names(getData(cObjMul, "features"))
  ),
  counts.log = getData(cObjSng, "counts.log"),
  classes = tibble(
    sample = colnames(getData(cObjSng, "counts")), 
    class = getData(cObjSng, "classification")
  )
)
```

Show top 10 (by fold change) features per class.

```{r, fig.align="center", fig.width=10, fig.height=8, eval = TRUE}
violinPlot <- function(cpm, genes, classes, class) {
  cpm[genes, ] %>%
    t() %>% 
    matrix_to_tibble("sample") %>%
    gather(gene, cpm, -sample) %>%
    full_join(tibble(sample = colnames(cpm), class = classes), by = "sample") %>%
    ggplot() +
    geom_jitter(aes(class, cpm), height = 0, size = 0.1, alpha = 0.25) +
    facet_wrap(~gene, scales = "free_y") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90)) +
    labs(title = paste0(class, " genes"))
}


getGenes <- function(
  cpm, classes, features, class, ntop = 10
) {
  fc <- foldChangePerClass(
    cpm[features[names(features) == class], ], 
    tibble(sample = colnames(cpm), class = classes)
  )
  rownames(fc[order(fc[, class], decreasing = TRUE), ])[1:ntop]
}

cpm <- getData(cObjSng, "counts.cpm")
c <- "C.Stem.distal"
g <- getGenes(cpm, getData(cObjSng, "classification"), getData(cObjMul, "features"), c)
violinPlot(cpm, g, getData(cObjSng, "classification"), c)

c <- "C.Stem.proximal"
g <- getGenes(cpm, getData(cObjSng, "classification"), getData(cObjMul, "features"), c)
violinPlot(cpm, g, getData(cObjSng, "classification"), c)

c <- "C.Goblet.distal"
g <- getGenes(cpm, getData(cObjSng, "classification"), getData(cObjMul, "features"), c)
violinPlot(cpm, g, getData(cObjSng, "classification"), c)

c <- "C.Goblet.proximal"
g <- getGenes(cpm, getData(cObjSng, "classification"), getData(cObjMul, "features"), c)
violinPlot(cpm, g, getData(cObjSng, "classification"), c)

c <- "SI.Stem"
g <- getGenes(cpm, getData(cObjSng, "classification"), getData(cObjMul, "features"), c)
violinPlot(cpm, g, getData(cObjSng, "classification"), c)

c <- "SI.Goblet"
g <- getGenes(cpm, getData(cObjSng, "classification"), getData(cObjMul, "features"), c)
violinPlot(cpm, g, getData(cObjSng, "classification"), c)

c <- "SI.Enterocytes"
g <- getGenes(cpm, getData(cObjSng, "classification"), getData(cObjMul, "features"), c)
violinPlot(cpm, g, getData(cObjSng, "classification"), c)

c <- "SI.Paneth"
g <- getGenes(cpm, getData(cObjSng, "classification"), getData(cObjMul, "features"), c)
violinPlot(cpm, g, getData(cObjSng, "classification"), c)
```

### Deconvolution

```{r, fig.align="center", fig.width=14, fig.height=12}
plotSwarmCircos(sObj, cObjSng, cObjMul, weightCut = 0, classOrder = cOrder)
```

### Filtering

Only detected duplicates and triplicates.  
Only ERCC estimated cell number <= 4.  
Weight cutoff = 5.

```{r, fig.align="center", fig.width=14, fig.height=12}
adj <- adjustFractions(cObjSng, cObjMul, sObj, binary = TRUE)
samples <- rownames(adj)
rs <- rowSums(adj)
keep <- rs == 2 | rs == 3

plotSwarmCircos(
  filterSwarm(sObj, keep), cObjSng, cObjMul, weightCut = 5, 
  classOrder = cOrder, theoretical.max = 4
)
```

#False positive stats

```{r}
efm <- getEdgesForMultiplet(sObj, cObjSng, cObjMul, theoretical.max = 4) %>% 
  mutate(conn = map2(from, to, ~tibble(
    from = sort(c(.x, .y))[1], to = sort(c(.x, .y))[2]
  ))) %>% 
  select(-from, -to) %>% 
  unnest() %>% 
  distinct()

edgeStats <- calculateEdgeStats(sObj, cObjSng, cObjMul, theoretical.max = 4) %>%
  mutate(conn = map2(from, to, ~tibble(
    from = sort(c(.x, .y))[1], to = sort(c(.x, .y))[2]
  ))) %>% 
  select(-from, -to) %>% 
  unnest() %>% 
  select(from, to, everything()) %>%
  distinct() %>%
  as_tibble()

fp <- efm %>%
  mutate(fp = case_when(
    str_detect(from, "^SI") & str_detect(to, "^C") ~ TRUE,
    str_detect(to, "^SI") & str_detect(from, "^C") ~ TRUE,
    str_detect(from, "proximal") & str_detect(to, "distal") ~ TRUE,
    str_detect(to, "proximal") & str_detect(from, "distal") ~ TRUE,
    TRUE ~ FALSE
  )) %>%
  group_by(sample) %>%
  summarize(fp = sum(fp)) %>%
  {setNames(pull(., fp), pull(., sample))}

fp.c <- efm %>%
  inner_join(select(MGA.Meta, sample, sub_tissue), by = "sample") %>%
  mutate(fp = case_when(
    (str_detect(from, "^C") | str_detect(to, "^C")) & sub_tissue == "small_intestine" ~ TRUE,
    (str_detect(from, "^SI") | str_detect(to, "^SI")) & sub_tissue == "colon" ~ TRUE,
    TRUE ~ FALSE
  )) %>%
  group_by(sample) %>%
  summarize(fp = sum(fp)) %>%
  {setNames(pull(., fp), pull(., sample))}

```

If we define a false positive as a connection between either, the small 
intestine and the colon or a connection between the proximal and distal colon 
based on the classifications, the number of detected false positives is as 
follows:  
(Definition 1) Detected `r sum(fp[fp > 0])` false positive connections out of 
`r nrow(efm)` total connections. Of those multiplets with 
a detected connection, `r length(fp[fp > 0])` multiplets have at least one false
positive out of `r length(fp)` total multiplets. The per multiplet false 
positive distribution is shown in the plot below.

```{r, fig.align="center", fig.width=14, fig.height=12}
ggplot() +
  geom_bar(aes(fp)) +
  theme_bw() +
  labs(x = "False positives per multiplet") +
  scale_x_continuous(breaks = 0:max(fp))
```

If we instead define the false positives as multiplets that were sorted as small
intestine that contain connections with colon classes and vice versa, the number 
of false positives is as follows:  
(Definition 2) Detected `r sum(fp.c[fp.c > 0])` false positive connections out of 
`r nrow(efm)` total connections. Of those multiplets with 
a detected connection, `r length(fp.c[fp.c > 0])` multiplets have at least one false
positive out of `r length(fp.c)` total multiplets.

We can also evaluate false positives via definition 1 and 2 and also include
whether or not the connection is significant at alpha 0.05.

(Definition 1)

```{r}
edgeStats %>% 
  mutate(significant = pval < 0.05) %>%
  mutate(`false positive` = case_when(
    str_detect(from, "^SI") & str_detect(to, "^C") ~ TRUE,
    str_detect(to, "^SI") & str_detect(from, "^C") ~ TRUE,
    str_detect(from, "proximal") & str_detect(to, "distal") ~ TRUE,
    str_detect(to, "proximal") & str_detect(from, "distal") ~ TRUE,
    TRUE ~ FALSE
  )) %>% 
  group_by(significant, `false positive`) %>% 
  summarize(n = sum(weight)) %>% 
  ungroup() %>% 
  as.data.frame()
```

(Definition 2)

```{r}

efm %>%
  inner_join(edgeStats, by = c("from", "to")) %>% 
  inner_join(select(MGA.Meta, sample, sub_tissue), by = "sample") %>%
  mutate(significant = pval < 0.05) %>%
  mutate(`false positive` = case_when(
    (str_detect(from, "^C") | str_detect(to, "^C")) & sub_tissue == "small_intestine" ~ TRUE,
    (str_detect(from, "^SI") | str_detect(to, "^SI")) & sub_tissue == "colon" ~ TRUE,
    TRUE ~ FALSE
  )) %>% 
  count(significant, `false positive`)
```

The type of false positive connections and their number are shown below.

(Definition 1)

```{r}
efm %>%
  mutate(fp = case_when(
    str_detect(from, "^SI") & str_detect(to, "^C") ~ TRUE,
    str_detect(to, "^SI") & str_detect(from, "^C") ~ TRUE,
    str_detect(from, "proximal") & str_detect(to, "distal") ~ TRUE,
    str_detect(to, "proximal") & str_detect(from, "distal") ~ TRUE,
    TRUE ~ FALSE
  )) %>%
  filter(fp) %>%
  count(from, to) %>%
  arrange(desc(n)) %>%
  as.data.frame()
```

(Definition 2)

```{r}
efm %>%
  inner_join(select(MGA.Meta, sample, sub_tissue)) %>%
  mutate(fp = case_when(
    (str_detect(from, "^C") | str_detect(to, "^C")) & sub_tissue == "small_intestine" ~ TRUE,
    (str_detect(from, "^SI") | str_detect(to, "^SI")) & sub_tissue == "colon" ~ TRUE,
    TRUE ~ FALSE
  )) %>%
  filter(fp) %>%
  count(from, to) %>%
  arrange(desc(n)) %>%
  as.data.frame()
```

Number of fp divided by number of detected connections.

```{r}
dc <- rowSums(adjustFractions(cObjSng, cObjMul, sObj))
tibble(
  sample = names(fp),
  `False positives (n)` = fp,
  `Deconvolution detected connections (n)` = dc[names(fp)]
) %>% 
  group_by(`Deconvolution detected connections (n)`) %>%
  summarize(`False positives (n)` = sum(`False positives (n)`)) %>%
  as.data.frame()
```

```{r, eval = FALSE, echo = FALSE}
sub_tissue_count <- tibble(sample = names(fp), fp = fp) %>% 
  inner_join(MGA.Meta, by = "sample") %>% 
  mutate(`False positive` = fp > 0) %>% 
  count(`False positive`, sub_tissue) %>%
  rename(`multiplets (n)` = n)

#counts the type of false positives per tissue
efm %>%
  mutate(`False positive type` = case_when(
    str_detect(from, "^SI") & str_detect(to, "^C") ~ "SC",
    str_detect(to, "^SI") & str_detect(from, "^C") ~ "SC",
    str_detect(from, "proximal") & str_detect(to, "distal") ~ "PD",
    str_detect(to, "proximal") & str_detect(from, "distal") ~ "PD",
    TRUE ~ "none"
  )) %>%
  inner_join(MGA.Meta) %>%
  count(sub_tissue, `False positive type`) %>%
  rename(`False positives (n)` = n)
```

Correlation between various parameters and the number of false positives is 
shown below:

```{r}
mulNames <- names(fp) #note other multiplets have 0 connections

ec <- estimateCells(cObjSng, cObjMul, warn = FALSE) %>%
  filter(sample %in% mulNames) %>%
  {setNames(pull(., estimatedCellNumber), pull(., sample))}
ec <- ec[mulNames]
dec <- rowSums(adjustFractions(cObjSng, cObjMul, sObj, binary = TRUE))
dec <- dec[mulNames]
c <- getData(sObj, "costs")
c <- c[mulNames]
tc <- colSums(getData(cObjMul, "counts"))
tc <- tc[mulNames]
dg <- apply(getData(cObjMul, "counts"), 2, function(c) length(which(c != 0)))
dg <- dg[mulNames]
feat.do <- getData(cObjMul, "counts.cpm")[features, ] %>%
  matrix_to_tibble("gene") %>%
  gather(sample, cpm, -gene) %>%
  group_by(sample) %>%
  summarize(do = length(which(cpm == 0))) %>%
  {setNames(pull(., do), pull(., sample))}
feat.do <- feat.do[mulNames]

cormat <- cor(matrix(c(fp, ec, dec, c, tc, dg, feat.do), ncol = 7))
corvec <- cormat[2:nrow(cormat), 1]
names(corvec) <- c(
  "ERCC estimated cell number", "Deconvolution estimated cell number", 
  "Deconvolution cost", "Total counts", "Total detected genes", 
  "Drop-outs in features"
)
data.frame(cor = corvec)
```

```{r, eval = FALSE}
cpm <- getData(cObjMul, "counts.cpm")
features <- getData(cObjMul, "features")
cpm[features, ] %>%
  matrix_to_tibble("gene") %>%
  gather(sample, cpm, -gene) %>%
  inner_join(tibble(class = names(features), gene = rownames(cpm)[features])) %>%
  group_by(sample, class) %>%
  summarize(cpm.sum = sum(cpm)) %>%
  ungroup() %>%
  inner_join(tibble(sample = names(fp), fp = fp)) %>%
  ggplot() +
  geom_point(aes(cpm.sum, fp)) +
  facet_wrap(~class)
#multiplets with high fp look to have low expression of the selected features

```

Distribution of fractions in connections thought to be true vs connections
known to be false.

```{r, fig.align="center", fig.width=14, fig.height=12}
fractions <- getData(sObj, "fractions")
efm %>%
  mutate(fp = case_when(
    str_detect(from, "^SI") & str_detect(to, "^C") ~ TRUE,
    str_detect(to, "^SI") & str_detect(from, "^C") ~ TRUE,
    str_detect(from, "proximal") & str_detect(to, "distal") ~ TRUE,
    str_detect(to, "proximal") & str_detect(from, "distal") ~ TRUE,
    TRUE ~ FALSE
  )) %>%
  mutate(fracs = pmap(list(sample, from, to), function(s, f, t) {
    fractions[s, c(f, t)]
  })) %>%
  unnest() %>%
  ggplot() +
  geom_boxplot(aes(fp, fracs)) +
  labs(x = "False positive", y = "Deconvolution fraction") +
  theme_bw()
```

Cluster multiplets and look for an overrepresentaion of false positives.

```{r, fig.align="center", fig.width=14, fig.height=12}
p.dist <- CIMseq::pearsonsDist(getData(cObjMul, "counts.cpm"), CIMseq::selectTopMax(getData(cObjMul, "counts.cpm"), 2000))
set.seed(234899)
u <- uwot::umap(p.dist)
rownames(u) <- colnames(getData(cObjMul, "counts"))
u %>% 
  matrix_to_tibble("sample") %>% 
  inner_join(tibble(sample = names(fp), fp = fp)) %>% 
  mutate(fp.bool = fp > 0) %>% 
  ggplot() + 
  geom_point(aes(V1, V2, colour = fp.bool)) +
  theme_bw() +
  labs(x = "UMAP1", y = "UMAP2") +
  guides(colour = guide_legend(title = "Has false positive?"))

u %>% 
  matrix_to_tibble("sample") %>% 
  inner_join(tibble(sample = names(fp), fp = fp)) %>% 
  ggplot() + 
  geom_point(aes(V1, V2, colour = fp)) +
  theme_bw() +
  scale_colour_viridis_c() +
  labs(x = "UMAP1", y = "UMAP2") +
  guides(colour = guide_legend(title = "False positives (n)"))
```

