---
title: "Validation of synthetic data algorithm"
author: "Jason T. Serviss"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r, message = FALSE}
#PACKAGES
packages <- c(
  "CIMseq","CIMseq.data", "CIMseq.testing", "printr", "ggthemes", "tibble",
  "stringr", "dplyr", "purrr", "tidyr", "ggplot2", "matrixStats", "readr"
)
purrr::walk(packages, library, character.only = TRUE)
rm(packages)

#FUNCTIONS
#cpm normalization without adding 1 that works on both matrices and tibbles with
#genes as rows and cells as columns
norm <- function(counts) {
  if(all(class(counts) == c("tbl_df", "tbl", "data.frame"))) {
    mat <- counts %>% 
      as.data.frame() %>%
      column_to_rownames("gene") %>%
      as.matrix()
    
    t(t(mat) / colSums(mat) * 10^6) %>%
      matrix_to_tibble("gene")
  } else if(class(counts) == "matrix") {
    t(t(counts) / colSums(counts) * 10^6)
  }
}

foldChangePerClass <- function(counts, classes) {
  uGroups <- unique(classes$class)
  
  res <- sapply(1:length(uGroups), function(x) {
    samplesA <- filter(classes, class == uGroups[x])$sample
    samplesB <- filter(classes, class != uGroups[x])$sample
    a <- rowMeans2(counts[, colnames(counts) %in% samplesA])
    b <- rowMeans2(counts[, colnames(counts) %in% samplesB])
    a / b
  })
  colnames(res) <- uGroups
  rownames(res) <- rownames(counts)
  return(res)
}
```

he sum of 
counts is calculated for each gene

The synthetic multiplets are based on the mixing of real synthetic singlets. The 
algorithm for generating the synthetic multiplets works as follows: Two singlets from 
each cell type contributing to the multiplet are randomly selected. The fraction of 
input into the multiplet can be adjusted so that, e.g. cell type A only contributes 
half of what cell type B contributes to the multiplet, by multiplying the singlet 
counts per million values with the desired fraction of input (fractions must sum to
1). The sum of counts per million is subsequently calculated for each gene providing
the counts per million of the synthesized multiplet.

Here we utilize the sorted cell line dataset to generate multiplets from the 
singlets. The synthesized multiplets are then compared to the real multiplets
to acertain the quality of the _in silico_ synthesis.

### Synthesize 10 multiplets per combination.

We begin by synthesizing ten multiplet per cell line combination, i.e. a 
A375-HCT116, A375-HOS, and HCT116-HOS multiplet.

```{r}
s <- str_detect(colnames(SCM.Counts), "^s")
sng <- SCM.Counts[, s]
classes <- slice(SCM.Meta, match(colnames(sng), SCM.Meta$sample))$cellTypes
n <- 10

fractions <- unique(classes) %>%
  sort() %>%
  combn(2) %>%
  as.data.frame(stringsAsFactors = FALSE) %>% 
  {paste(.[1, ], .[2, ], sep = "-")} %>%
  c(., "A375-HCT116-HOS") %>%
  setNames(., .) %>%
  map(function(x) {
    vec <- rep(0, length(unique(classes)))
    names(vec) <- unique(classes)
    sp <- str_split(x, "-")[[1]]
    vec[names(vec) %in% sp] <- 1/length(sp)
    vec[!names(vec) %in% sp] <- (1 - sum(vec)) / length(which(vec == 0))
    vec
  }) %>%
  rep(., n)


synth <- map(1:length(fractions), function(i) {
    syntheticMultipletsFromCounts(norm(sng), classes, fractions[[i]], seed = 92830 + i)
  }) %>%
  do.call("cbind", .) %>%
  matrix_to_tibble("gene") %>%
  mutate(gene = rownames(SCM.Counts)) %>%
  setNames(c(
    "gene",
    paste(names(fractions), rep(1:n, each = length(fractions) / n), sep = ".")
  )) %>%
  gather(cellTypes, `Synthetic counts`, -gene) %>%
  separate(cellTypes, into = c("cellTypes", "syntheticMultipletID"), sep = "\\.")

mul <- SCM.Counts[, !s] %>%
  norm() %>%
  matrix_to_tibble("gene") %>%
  gather(sample, `Counts`, -gene) %>%
  left_join(SCM.Meta, by = "sample") %>%
  select(gene, `Counts`, cellTypes, sample)
```

Plot results.

```{r, fig.align="center", fig.width=10, fig.height=8, message = FALSE, eval = FALSE}
p1 <- synth %>%
  inner_join(mul, by = c("gene", "cellTypes")) %>%
  gather(type, value, -(gene:syntheticMultipletID), -sample) %>%
  filter(cellTypes != "A375-HCT116-HOS") %>%
  mutate(type = if_else(type == "Counts", "Real multiplet CPM", "Synthetic multiplet CPM")) %>%
  ggplot() +
  geom_histogram(
    aes(log2(value + 1), stat(ndensity), fill = type),
    alpha = 0.5, binwidth = 0.25, position = "identity"
  ) +
  facet_wrap(~cellTypes) +
  labs(x = "log2(cpm + 1)", y = "Density") +
  theme_bw() +
  ggthemes::scale_fill_ptol() +
  guides(fill = guide_legend(title = "")) +
  theme(legend.position = "top")

p1

ggsave(
  plot = p1,
  filename = '../figures/figure1.pdf',
  device = cairo_pdf,
  height = 180,
  width = 240,
  units = "mm"
)
```

```{r, fig.align="center", fig.width=10, fig.height=8, message = FALSE}
p1 <- synth %>%
    inner_join(mul, by = c("gene", "cellTypes")) %>%
    gather(type, value, -(gene:syntheticMultipletID), -sample) %>%
    filter(cellTypes != "A375-HCT116-HOS") %>%
    mutate(type = if_else(type == "Counts", "Real multiplet", "Synthetic multiplet")) %>%
    rename(cpm = value) %>%
    ggplot() +
    geom_violin(
      data = . %>% filter(cpm != 0),
      aes(cellTypes, log2(cpm + 1), fill = type), alpha = 0.5, position = position_identity()
    ) +
    geom_text(
      data = . %>%
      group_by(cellTypes, type) %>%
      summarize(do = paste0(round(length(which(cpm == 0)) / n(), digits = 2) * 100, "%")),
      aes(cellTypes, 0.05, label = do, group = type, colour = type), 
      size = 5, position = position_dodge(width = 0.9)
    ) +
    scale_colour_ptol() +
    scale_fill_ptol() +
    labs(x = "Multiplet composition") +
    theme_few() +
    theme(legend.position = "top", legend.justification = "center") +
    guides(fill = guide_legend(title = ""), colour = FALSE)

p1

p2 <- synth %>%
    inner_join(mul, by = c("gene", "cellTypes")) %>%
    gather(type, value, -(gene:syntheticMultipletID), -sample) %>%
    filter(cellTypes != "A375-HCT116-HOS") %>%
    mutate(type = if_else(type == "Counts", "Real multiplet", "Synthetic multiplet")) %>%
    rename(cpm = value) %>%
    ggplot() +
    geom_violin(
      data = . %>% filter(cpm != 0),
      aes("x", log2(cpm + 1), fill = type), alpha = 0.5, position = position_identity()
    ) +
    geom_text(
      data = . %>%
      group_by(type) %>%
      summarize(do = paste0(round(length(which(cpm == 0)) / n(), digits = 2) * 100, "%")),
      aes("x", 0.05, label = do, group = type, colour = type), 
      size = 5, position = position_dodge(width = 0.9)
    ) +
    scale_colour_ptol() +
    scale_fill_ptol() +
    theme_few() +
    theme(
      legend.position = "top", 
      legend.justification = "center",
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title.x = element_blank()
    ) +
    guides(fill = guide_legend(title = "", nrow = 2), colour = FALSE)

p2

ggsave(
  plot = p2,
  filename = '../figures/figure1.pdf',
  device = cairo_pdf,
  height = 180,
  width = 50,
  units = "mm"
)
```

<br></br>

### Adjusting multiplet contribution

Synthesize 25 multiplets for each combination and reduce the contribution
of HOS cells by 0%, 25%, 50%, 75%, or 99%. Calculate the mean gene expression of
25 cell type specific genes for each combination and plot together with the real
multiplets and corresponding singlets.

Show genes used in figure.
```{r}
#find overexpressed genes in each cell type
genes <- SCM.Meta %>%
  filter(cellNumber == "Singlet") %>%
  rename(class = cellTypes) %>%
  foldChangePerClass(CIMseq:::.norm.counts(sng), .) %>%
  matrix_to_tibble("gene") %>%
  select(gene, A375, HCT116, HOS) %>%
  gather(geneType, fold, -gene) %>%
  filter(!is.na(fold) & !is.infinite(fold)) %>%
  group_by(geneType) %>%
  top_n(25, fold) %>%
  ungroup()

as.data.frame(genes)
```

Plot figure.

```{r}
###FUNCTIONS
#function to calculate the mean gene expression for a specific cell type. 
typeMeans <- function(type) {
  SCM.Meta %>%
    filter(cellTypes == type) %>% 
    pull(sample) %>%
    {SCM.Counts[, colnames(SCM.Counts) %in% .]} %>%
    norm() %>%
    rowMeans2() %>%
    tibble(gene = rownames(SCM.Counts), counts = .) %>%
    mutate(cellType = type)
}

#Synthesize 25 multiplets with varying levels of adjusted HOS input
#Calculate the mean gene expression in the 25 multiplets for each condition
s <- str_detect(colnames(SCM.Counts), "^s")
sng <- SCM.Counts[, s]
classes <- slice(SCM.Meta, match(colnames(sng), SCM.Meta$sample))$cellTypes

#setup fractions
.assignFractions <- function(ct, adj) {
  vec <- rep(0, length(unique(classes)))
  names(vec) <- sort(unique(classes))
  if("HOS" %in% ct) {
    vec[names(vec) == "HOS"] <- 1/length(ct) * adj
    vec[names(vec) %in% ct[ct != "HOS"]] <- (1 - sum(vec)) / length(ct[ct != "HOS"])
  } else {
    vec[names(vec) %in% ct] <- 1/length(ct)
    vec[!names(vec) %in% ct] <- (1 - sum(vec)) / length(which(vec == 0))
  }
  vec
}

adj <- c(1, 0.75, 0.5, 0.25, 0.01)
n <- 25

combos <- unique(classes) %>%
  sort() %>%
  combn(2) %>%
  as.data.frame(stringsAsFactors = FALSE) %>% 
  {paste(.[1, ], .[2, ], sep = "-")}

set.seed(195)
fractions <- tibble(
  cellType = rep(combos, each = length(adj) * n),
  adjustment = rep(adj, length(combos) * n),
  ns = rep(rep(1:n, each = length(adj), length(combos)))
) %>%
  mutate(cellTypes = map(cellType, function(x) str_split(x, "-")[[1]])) %>%
  mutate(fractions = map2(cellTypes, adjustment, ~.assignFractions(.x, .y))) %>%
  mutate(seed = runif(1, 1, 99999) + 1:n())

#synthesize multiplets
synth <- fractions %>%
  mutate(sm = map2(fractions, seed, function(f, s) {
    syntheticMultipletsFromCounts(norm(sng), classes, f, seed = s) %>%
      matrix_to_tibble("gene") %>%
      mutate(gene = rownames(SCM.Counts)) #HERE
  })) %>% 
  select(cellType, adjustment, ns, sm) %>%
  unnest() %>%
  group_by(cellType, adjustment, gene) %>%
  summarize(counts_synthetic = mean(V1)) %>%
  ungroup() %>%
  spread(adjustment, counts_synthetic) %>%
  setNames(c(
    "cellType", "gene", 
    paste0("HOS reduced ", (1 - as.numeric(colnames(.)[-c(1:2)])) * 100, "%")
  ))

#calculate the mean gene expression for each type of singlet
realSng <- map_dfr(c("A375", "HCT116", "HOS"), typeMeans)

#calculate the mean gene expression for each type of multiplet
realMul <- map_dfr(c("A375-HCT116", "A375-HOS", "HCT116-HOS"), typeMeans) %>%
  rename(Multiplet = counts) %>% 
  separate(cellType, into = c("cellType1", "cellType2"), sep = "-", remove = FALSE)

#concatenate real singlet, real multiplet, and info for genes corresponding to 
#a cell type
realData <- realMul %>%
  inner_join(realSng, by = c("cellType1" = "cellType", "gene" = "gene")) %>% 
  inner_join(realSng, by = c("cellType2" = "cellType", "gene" = "gene")) %>%
  rename(counts_cellType1 = counts.x, counts_cellType2 = counts.y) %>%
  inner_join(genes, by = "gene") %>%
  filter(geneType == cellType1 | geneType == cellType2) %>%
  select(
    gene, geneType, cellType, cellType1, cellType2, 
    counts_cellType1, counts_cellType2, Multiplet
  )

#concatenate real and synthetic data and prepare for plotting
plotData <- realData %>% 
  inner_join(synth, by = c("gene", "cellType")) %>%
  gather(cpmType, cpm, -gene, -geneType, -cellType, -cellType1, -cellType2) %>%
  mutate(cpmType = case_when(
    cpmType == "counts_cellType1" ~ paste0(cellType1, " singlets"),
    cpmType == "counts_cellType2" ~ paste0(cellType2, " singlets"),
    cpmType == "Multiplet" ~ paste0(cellType, " multiplets"),
    TRUE ~ cpmType
  )) %>%
  mutate(cpmType = parse_factor(cpmType, levels = c(
      "A375-HOS multiplets", "HCT116-HOS multiplets", "A375-HCT116 multiplets", 
      "HOS reduced 0%", "HOS reduced 25%", "HOS reduced 50%", "HOS reduced 75%", 
      "HOS reduced 99%", "A375 singlets", "HCT116 singlets", "HOS singlets"
  ))) %>%
  mutate(plotGene = paste0(gene, " (", geneType, ")")) %>%
  mutate(plotGene = parse_factor(plotGene, levels = unique(plotGene))) %>%
  mutate(cellType = parse_factor(cellType, levels = c("A375-HOS", "HCT116-HOS", "A375-HCT116", "A375-HCT116-HOS")))
```

```{r, fig.align="center", fig.width=10, fig.height=8, message = FALSE}
p3 <- plotData %>%
  ggplot() +
  geom_line(aes(cpmType, log2(cpm + 1), group = gene, colour = geneType)) +
  geom_point(aes(cpmType, log2(cpm + 1), colour = geneType), size = 0.75) +
  facet_wrap(~cellType, scales = "free") +
  theme_few() +
  ggthemes::scale_colour_ptol() +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_text(angle = 90),
    legend.position = "top"
  ) +
  guides(colour = guide_legend(
    title = "Gene's cell type specificity",
    title.position = "top", 
    title.hjust = 0.5)
  )

p3

ggsave(
  plot = p3,
  filename = '../figures/figure2.pdf',
  device = cairo_pdf,
  height = 180,
  width = 240,
  units = "mm"
)
```

```{r, fig.align="center", fig.width=10, fig.height=8, message = FALSE, eval = FALSE}


p3 <- plotData %>%
  group_by(cpmType, cellType, geneType) %>%
  summarize(m = mean(cpm)) %>%
  ggplot() +
  geom_line(aes(cpmType, log2(m + 1), group = geneType, colour = geneType)) +
  geom_point(aes(cpmType, log2(m + 1), colour = geneType), size = 0.75) +
  facet_wrap(~cellType, scales = "free") +
  theme_bw() +
  ggthemes::scale_colour_ptol() +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_text(angle = 90),
    legend.position = "top"
  ) +
  guides(colour = guide_legend(
    title = "Gene's cell type specificity",
    title.position = "top", 
    title.hjust = 0.5)
  )

p3

ggsave(
  plot = p3,
  filename = '../figures/figure2.pdf',
  device = cairo_pdf,
  height = 180,
  width = 240,
  units = "mm"
)
```

The results show a minimal change in gene expression between the sorted 
multiplets and the non-adjusted (0%) multiplets (first two columns). In 
addition, the specificity of the genes for the individual cell types is easily
recognized (last two columns). Finally, it can be seen that decreasing levels
of HOS input into the synthetic multiplets causes a concomitatnt decrease in 
HOS specific gene expression in synthetic multiplets including the HOS cell 
type. On the other hand, in multiplets that do not include HOS cells, the gene 
expression of the HOS genes cause no changes in gene expression.

In summary, the results indicate that the algroithm for generating synthetic 
multiplets is capable of mirroring the gene expression profile in multiplets
of a known composition. Furthermore, adjusting the input of a specific cell type
in the multiplet results in the expected changes in the multiplets gene 
expression.

<br></br>

