---
title: "sortedMultiplets20171116-distToSlice"
author: "Jason T. Serviss"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r, echo = FALSE}
packages <- c(
    "sp.scRNAseq",
    "sp.scRNAseqTesting",
    "sp.scRNAseqData",
    "printr",
    "ggthemes",
    "tidyverse"
)
purrr::walk(packages, library, character.only = TRUE)
rm(packages)
```

Run method.
```{r}
sng <- str_detect(colnames(countsSorted2), "^s")

#create counts objects
cObjSng <- spCounts(countsSorted2[, sng], countsSortedERCC2[, sng])
cObjMul <- spCounts(countsSorted2[, !sng], countsSortedERCC2[, !sng])

#spUnsupervised
uObj <- spUnsupervised(cObjSng)

#rename classes
positions <- str_extract(colnames(getData(cObjSng, "counts")), "...$")
newClass <- case_when(
  positions == "E03" ~ "A375", #what is this? sorting issue?
  positions %in% paste0(sort(rep(LETTERS[1:8], 4)), c("01", "02", "03", "04")) ~ "HOS",
  positions %in% paste0(sort(rep(LETTERS[1:8], 4)), c("05", "06", "07", "08")) ~ "HCT116",
  positions %in% paste0(sort(rep(LETTERS[1:8], 4)), c("09", "10", "11", "12")) ~ "A375",
  TRUE ~ "error"
)
corresp <- getData(uObj, "classification") %>%
  tibble(oldClass = ., newClass = newClass) %>%
  distinct()

classification(uObj) <- newClass

gm <- getData(uObj, "groupMeans")
colnames(gm) <- pull(corresp, newClass)[match(colnames(gm), pull(corresp, oldClass))]
groupMeans(uObj) <- gm

tm <- getData(uObj, "tsneMeans")
tm$classification <- pull(corresp, newClass)[match(tm$classification, pull(corresp, oldClass))]
tsneMeans(uObj) <- tm

#spSwarm
sObj_distToSlice <- spSwarm(cObjMul, uObj, distFun = "distToSlice", swarmsize = 50, maxiter = 10)
sObj_bic <- spSwarm(cObjMul, uObj, distFun = "bic", swarmsize = 50, maxiter = 10)
#sObj_distToSliceNorm <- spSwarm(cObjMul, uObj, distFun = "distToSliceNorm", swarmsize = 50, maxiter = 10)
#sObj_distToSliceTop <- spSwarm(cObjMul, uObj, distFun = "distToSliceTop", swarmsize = 50, maxiter = 10, cells = 4)
#sObj_distToSliceEuclid <- spSwarm(cObjMul, uObj, distFun = "distToSliceEuclid", swarmsize = 50, maxiter = 10)
#sObj_distToSlicePearson <- spSwarm(cObjMul, uObj, distFun = "distToSlicePearson", swarmsize = 50, maxiter = 10)
```

Evaluate the information still contained in the permuted data. Do this by randomly selecting 5000 genes from the matrix and running pca

Use class means, calculated from the permuted counts matrix, to calculate the 
```{r, fig.align='center', fig.height=8, fig.width=10}

#calculate costs
cost_distToSlice <- permutationCosts(
  groupMeanPermutations,
  cObjMul,
  uObj,
  sObj_distToSlice,
  100,
  distFun = sp.scRNAseq:::distToSlice
)

cost_bic <- permutationCosts(
  groupMeanPermutations,
  cObjMul,
  uObj,
  sObj_bic,
  100,
  distFun = sp.scRNAseq:::bic
)

# cost_distToSliceNorm <- permutationCosts(
#   groupMeanPermutations,
#   cObjMul, 
#   uObj,
#   sObj_distToSliceNorm,
#   100,
#   distFun = sp.scRNAseq:::distToSliceNorm
# )

#cost_distToSliceTop <- permutationCosts(
#  groupMeanPermutations,
#  cObjMul, 
#  uObj, 
#  sObj_distToSliceTop, 
#  100, 
#  distFun = sp.scRNAseq:::distToSliceTop
#)

# cost_distToSliceEuclid <- permutationCosts(
#   groupMeanPermutations,
#   cObjMul, 
#   uObj, 
#   sObj_distToSliceEuclid,
#   100,
#   distFun = sp.scRNAseq:::distToSliceEuclid
# )

# cost_distToSlicePearson <- permutationCosts(
#   groupMeanPermutations,
#   cObjMul,
#   uObj,
#   sObj_distToSlicePearson,
#   100,
#   distFun = sp.scRNAseq:::distToSlicePearson
# )
```

Plot results for distToSlice.
```{r, fig.align='center', fig.height=8, fig.width=10}
plotDistPermutations(cost_distToSlice)
```


```{r, fig.align='center', fig.height=8, fig.width=10, eval = FALSE, echo = FALSE}
#Plot results for distToSliceNorm.
plotDistPermutations(cost_distToSliceNorm)
```

```{r, fig.align='center', fig.height=8, fig.width=10, eval = FALSE, echo = FALSE}
#Plot results for distToSliceTop.
#plotDistPermutations(cost_distToSliceTop)
```

```{r, fig.align='center', fig.height=8, fig.width=10, eval = FALSE, echo = FALSE}
#Plot results for distToSliceEuclid.
plotDistPermutations(cost_distToSliceEuclid)
```


```{r, fig.align='center', fig.height=8, fig.width=10, eval = FALSE, echo = FALSE}
#Plot results for distToSlicePearson.
plotDistPermutations(cost_distToSlicePearson)
```

Plot results for bic.
```{r, fig.align='center', fig.height=8, fig.width=10}
plotDistPermutations(cost_bic)
```

###Plot for individual multiplets

Plot results for distToSlice.
```{r, fig.align='center', fig.height=8, fig.width=10}
plotDistPermutations(cost_distToSlice, TRUE)
```

Plot results for distToSliceNorm.
```{r, fig.align='center', fig.height=8, fig.width=10}
plotDistPermutations(cost_distToSliceNorm, TRUE)
```

Plot results for distToSliceTop.
```{r, fig.align='center', fig.height=8, fig.width=10}
#plotDistPermutations(cost_distToSliceTop, TRUE)
```

Plot results for distToSliceEuclid.
```{r, fig.align='center', fig.height=8, fig.width=10}
plotDistPermutations(cost_distToSliceEuclid, TRUE)
```

Plot results for distToSlicePearson.
```{r, fig.align='center', fig.height=8, fig.width=10}
plotDistPermutations(cost_distToSlicePearson, TRUE)
```

Plot results for bic.
```{r, fig.align='center', fig.height=8, fig.width=10}
plotDistPermutations(cost_bic, TRUE)
```
