---
title: "Mouse colon analysis, 20 classes"
author: "Jason T. Serviss"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r, message = FALSE}
packages <- c("CIMseq", "CIMseq.data", "tidyverse", "circlize", "printr")
purrr::walk(packages, library, character.only = TRUE)
rm(packages)

##DATA
load('../data/CIMseqData.rda')
load('../data/sObj.rda')

#there are 5 cells that were classified as colon but sorted as SI. These have to
#be removed manually
c <- getData(cObjSng, "classification")
s <- names(c[c %in% c("4", "9")])
i <- which(colnames(getData(cObjSng, "counts")) %in% s)
cObjSng <- CIMseqSinglets(
  getData(cObjSng, "counts")[, -i],
  getData(cObjSng, "counts.ercc")[, -i],
  getData(cObjSng, "dim.red")[-i, ],
  getData(cObjSng, "classification")[-i]
)
```

```{r, eval = TRUE, echo = FALSE}
#rename classes
renameClasses <- function(class) {
  case_when(
    class == "0" ~ "C.Lgr5+.proximal.2",
    class == "1" ~ "C.Goblet.distal.1",
    class == "2" ~ "C.Colonocytes",
    class == "5" ~ "C.Lgr5+.distal",
    class == "6" ~ "C.TA.distal",
    class == "7" ~ "C.Lgr5+.Mki67+",
    class == "8" ~ "C.Goblet.proximal",
    class == "10" ~ "C.Goblet.distal.Plet1",
    class == "13" ~ "C.Lgr5+.proximal.1",
    class == "14" ~ "Enteroendocrine",
    class == "15" ~ "Tufft",
    class == "18" ~ "C.Goblet.Mki67+",
    class == "19" ~ "Blood",
    TRUE ~ "error"
  )
}

cOrder <- c(
  "C.Goblet.proximal", "C.Lgr5+.proximal.1", "C.Lgr5+.proximal.2", "C.Colonocytes", "C.Lgr5+.Mki67+",
  "C.Goblet.Mki67+", "C.Goblet.distal.Plet1", "C.Goblet.distal.1", "C.Lgr5+.distal", "C.TA.distal",
  "Enteroendocrine", "Tufft", "Blood"
)

getData(cObjSng, "classification") <- renameClasses(getData(cObjSng, "classification"))
fractions <- getData(sObj, "fractions")
fractions <- fractions[, !colnames(fractions) %in% c("4", "9")]
colnames(fractions) <- renameClasses(colnames(fractions))
sObj@fractions <- fractions
```


### Fig 1: classes
```{r, fig.align="center", fig.width=10, fig.height=8}
p <- plotUnsupervisedClass(cObjSng, cObjMul)
p
ggsave(
  plot = p,
  filename = '../figures/MGA.enge20.classes.pdf',
  device = cairo_pdf,
  height = 180,
  width = 180,
  units = "mm"
)

```

### Fig 2: Cell type gene expression
```{r, fig.align="center", fig.width=10, fig.height=8}
p <- plotUnsupervisedMarkers(
  cObjSng, cObjMul,
  c("Lgr5", "Ptprc", "Chga", "Dclk1", "Slc26a3", "Atoh1"),
  pal = RColorBrewer::brewer.pal(8, "Set1")
)
ggsave(
  plot = p,
  filename = '../figures/MGA.enge20.markers.pdf',
  device = cairo_pdf,
  height = 180,
  width = 180,
  units = "mm"
)

```

### Fig 3: Cell cycle
```{r, fig.align="center", fig.width=10, fig.height=8}
p <- plotUnsupervisedMarkers(
  cObjSng, cObjMul, c("Mki67"),
  pal = RColorBrewer::brewer.pal(8, "Set1")
)
ggsave(
  plot = p,
  filename = '../figures/MGA.enge20.Mki67.pdf',
  device = cairo_pdf,
  height = 180,
  width = 180,
  units = "mm"
)
```

### Fig 4: Connections per multiplet
```{r}
adj <- adjustFractions(cObjSng, cObjMul, sObj)
as.data.frame(table(apply(adj, 1, sum)))
```

### Fig 5: Fraction histogram
```{r, fig.align="center", fig.width=10, fig.height=8}
tibble(fractions = c(fractions)) %>%
  ggplot() +
  geom_histogram(aes(fractions), binwidth = 0.01) +
  theme_bw()
```

### Fig 6: Detected cell types vs. cost
```{r, fig.align="center", fig.width=10, fig.height=8}
tibble(
  nCellTypes = apply(adj, 1, sum),
  cost = getData(sObj, "costs")
) %>%
  ggplot() +
  geom_boxplot(aes(nCellTypes, cost, group = nCellTypes)) +
  scale_x_continuous(name = "Detected cell types", breaks = 0:max(apply(adj, 1, sum))) +
  theme_bw()
```

### Fig 7: Estimated cell numbers vs. cost
```{r, fig.align="center", fig.width=10, fig.height=8}
tibble(
  sample = names(getData(sObj, "costs")),
  cost = unname(getData(sObj, "costs"))
) %>%
  inner_join(
    select(estimateCells(cObjSng, cObjMul), sample, estimatedCellNumber), 
    by = "sample"
  ) %>%
  mutate(estimatedCellNumber = round(estimatedCellNumber)) %>%
  ggplot() +
  geom_boxplot(aes(estimatedCellNumber, cost, group = estimatedCellNumber)) +
  scale_x_continuous(
    name = "ERCC estimated cell number", 
    breaks = 0:max(round(pull(estimateCells(cObjSng, cObjMul), estimatedCellNumber)))
  ) +
  theme_bw()
```

### Fig 8: Estimated cell number vs. Detected cell number
```{r, fig.align="center", fig.width=10, fig.height=8}
ercc <- filter(estimateCells(cObjSng, cObjMul), sampleType == "Multiplet")
nConnections <- apply(adj, 1, sum)
nConnections <- nConnections[match(ercc$sample, names(nConnections))]
tibble(
  detectedConnections = round(nConnections),
  estimatedCellNumber = round(ercc$estimatedCellNumber)
) %>%
  ggplot() +
  geom_boxplot(aes(estimatedCellNumber, detectedConnections, group = estimatedCellNumber)) +
  scale_x_continuous(
    name = "ERCC estimated cell number", 
    breaks = 0:max(round(ercc$estimatedCellNumber))
  ) +
  scale_y_continuous(
    name = "Detected cell number",
    breaks = 0:max(round(nConnections))
  ) +
  theme_bw()
```

### Fig 9: Detected cell number vs. Total counts
```{r, fig.align="center", fig.width=10, fig.height=8}
tibble(
  sample = names(nConnections),
  detectedConnections = nConnections
) %>%
  inner_join(tibble(
    sample = colnames(getData(cObjMul, "counts")),
    total.counts = colSums(getData(cObjMul, "counts"))
  ), by = "sample") %>%
  ggplot() +
  geom_boxplot(aes(detectedConnections, total.counts, group = detectedConnections)) +
  scale_x_continuous(
    name = "Detected cell number", 
    breaks = 0:max(nConnections)
  ) +
  scale_y_continuous(name = "Total counts") +
  theme_bw()
```

### Fig 10: Detected cell number vs. Total ERCC counts
```{r, fig.align="center", fig.width=10, fig.height=8}
tibble(
  sample = names(nConnections),
  detectedConnections = nConnections
) %>%
  inner_join(tibble(
    sample = colnames(getData(cObjMul, "counts")),
    total.ercc = colSums(getData(cObjMul, "counts.ercc"))
  ), by = "sample") %>%
  ggplot() +
  geom_boxplot(aes(detectedConnections, total.ercc, group = detectedConnections)) +
  scale_x_continuous(
    name = "Detected cell number", 
    breaks = 0:max(nConnections)
  ) +
  scale_y_continuous(name = "Total ERCC counts") +
  theme_bw()
```

### Fig 11: Relative frequency of singlets vs. deconvoluted multiplets

```{r, fig.align="center", fig.width=10, fig.height=8}
singlets <- c(table(getData(cObjSng, "classification")))
singlets <- singlets / sum(singlets)
deconv <- colSums(adjustFractions(cObjSng, cObjMul, sObj))
deconv <- deconv[match(names(singlets), names(deconv))]
deconv <- deconv / sum(deconv)
if(!identical(names(singlets), names(deconv))) stop("name mismatch")

p <- tibble(
  class = names(singlets),
  singlet.freq = singlets,
  multiplet.freq = deconv
) %>%
  ggplot() +
  geom_point(aes(singlet.freq, multiplet.freq, colour = class), size = 3) +
  scale_colour_manual(values = col40()) +
  xlim(min(c(deconv, singlets)), max(c(deconv, singlets))) +
  ylim(min(c(deconv, singlets)), max(c(deconv, singlets))) +
  geom_abline(slope = 1, intercept = 0, lty = 3, colour = "grey") +
  labs(x = "Singlet relative frequency", y = "Multiplet relative frequency") +
  guides(colour = guide_legend(title = "Cell Type")) +
  theme_bw()

p
ggsave(
  plot = p,
  filename = '../figures/MGA.enge20.sngMulRelFreq.pdf',
  device = cairo_pdf,
  height = 180,
  width = 180,
  units = "mm"
)
```

### Fig 12: All connections

```{r, fig.align="center", fig.width=10, fig.height=8}
plotSwarmCircos(sObj, cObjSng, cObjMul, classOrder = cOrder)
```

### Fig 13: Filtered

Only detected duplicates, triplicates, and quadruplicates.  
ERCC estimated cell number set to max 4.  
Weight cutoff = 10.

```{r, fig.align="center", fig.width=14, fig.height=12}
adj <- adjustFractions(cObjSng, cObjMul, sObj, binary = TRUE)
samples <- rownames(adj)
rs <- rowSums(adj)
keep <- rs == 2 | rs == 3 | rs == 4

plotSwarmCircos(
  filterSwarm(sObj, keep), cObjSng, cObjMul, weightCut = 10, 
  classOrder = cOrder, theoretical.max = 4
)

pdf('../figures/MGA.enge20.circos.pdf', width = 9.5, height = 9.5)
plotSwarmCircos(
  filterSwarm(sObj, keep), cObjSng, cObjMul, weightCut = 10, 
  classOrder = cOrder, theoretical.max = 4
)
dev.off()
```

