---
title: "Examining the accuracy of the fractions reported by spSwarm"
author: "Jason T. Serviss"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r, message = FALSE}
packages <- c(
    "sp.scRNAseq",
    "sp.scRNAseqData",
    "sp.scRNAseqTesting",
    "printr",
    "ggthemes",
    "tidyverse"
)
purrr::walk(packages, library, character.only = TRUE)
rm(packages)
```

```{r, message = FALSE}
#setup information
#order everything by column
multiplets <- c(
  c(rep("A375-HCT116", 3), rep("HCT116-HOS", 2), rep("A375-HOS", 3)),
  c(rep("A375-HCT116", 3), rep("HCT116-HOS", 2), rep("A375-HOS", 3)),
  c(rep("A375-HCT116", 2), rep("HCT116-HOS", 3), rep("A375-HOS", 3)),
  c(rep("A375-HCT116", 2), rep("HCT116-HOS", 3), rep("A375-HOS", 3)),
  c(rep("A375-HCT116-HOS", 3), rep("HCT116-HCT116-HOS", 2), rep("A375-HOS-HOS", 3)),
  c(rep("A375-HCT116-HOS", 3), rep("HCT116-HCT116-HOS", 2), rep("A375-HOS-HOS", 3)),
  c(rep("A375-HCT116-HOS", 2), rep("HCT116-HCT116-HOS", 3), rep("A375-HOS-HOS", 3)),
  c(rep("A375-HCT116-HOS", 2), rep("HCT116-HCT116-HOS", 3), rep("A375-HOS-HOS", 3)),
  c(rep("A375-HOS-HOS-HOS", 3), rep("A375-A375-HCT116-HCT116", 2), rep("A375-A375-HCT116-HOS", 3)),
  c(rep("A375-HOS-HOS-HOS", 3), rep("A375-A375-HCT116-HCT116", 2), rep("A375-A375-HCT116-HOS", 3)),
  c(rep("A375-HOS-HOS-HOS", 2), rep("A375-A375-HCT116-HCT116", 3), rep("A375-A375-HCT116-HOS", 3)),
  c(rep("A375-HOS-HOS-HOS", 2), rep("A375-A375-HCT116-HCT116", 3), rep("A375-A375-HCT116-HOS", 3))
)

cols <- c("01", "02", "03", "04", "05", "06", "07", "08", "09", "10", "11", "12")
rows <- LETTERS[1:8]
names <- paste0("m.NJB00204.", rep(rows, 12), sort(rep(cols, 8)))

#adds an ID indicating multiplets that should be exactly the same
addID <- function(data) {
  data %>%
    mutate(id = case_when(
    (row %in% c("A", "B") & column %in% c("01", "02", "03", "04")) | (row == "C" & column %in% c("01", "02")) ~ 1,
    (row %in% c("D", "E") & column %in% c("01", "02", "03", "04")) | (row == "C" & column %in% c("03", "04")) ~ 2,
    row %in% c("F", "G", "H") & column %in% c("01", "02", "03", "04") ~ 3,
    (row %in% c("A", "B") & column %in% c("05", "06", "07", "08")) | (row == "C" & column %in% c("05", "06")) ~ 4,
    (row %in% c("D", "E") & column %in% c("05", "06", "07", "08")) | (row == "C" & column %in% c("07", "08")) ~ 5,
    row %in% c("F", "G", "H") & column %in% c("05", "06", "07", "08") ~ 6,
    (row %in% c("A", "B") & column %in% c("09", "10", "11", "12")) | (row == "C" & column %in% c("09", "10")) ~ 7,
    (row %in% c("D", "E") & column %in% c("09", "10", "11", "12")) | (row == "C" & column %in% c("11", "12")) ~ 8,
    row %in% c("F", "G", "H") & column %in% c("09", "10", "11", "12") ~ 9
  ))
}

#adds the row and column data
addRowAndColumn <- function(data) {
  data %>%
    mutate(cellsInWell = case_when(
      str_extract(multiplet, "..$") %in% c("01", "02", "03", "04") ~ 2L,
      str_extract(multiplet, "..$") %in% c("05", "06", "07", "08") ~ 3L,
      str_extract(multiplet, "..$") %in% c("09", "10", "11", "12") ~ 4L
  ))
}

#Function to add the expected fractions. Fraction order is HOS, HCT116, A375
addExpectedFrac <- function(id) {
  case_when(
    id == 1 ~ c(0, 0.5, 0.5),
    id == 2 ~ c(0.5, 0.5, 0),
    id == 3 ~ c(0.5, 0, 0.5),
    id == 4 ~ c(1/3, 1/3, 1/3),
    id == 5 ~ c(1/3, 2/3, 0),
    id == 6 ~ c(2/3, 0, 1/3),
    id == 7 ~ c(3/4, 0, 1/4),
    id == 8 ~ c(0, 0.5, 0.5),
    id == 9 ~ c(1/4, 1/4, 0.5)
  )
}

#make plate data
plateData <- tibble(
  row = rep(rows, 12),
  column = sort(rep(cols, 8)),
  multipletName = names,
  multipletComposition = multiplets
) %>%
  mutate(connections = str_split(multipletComposition, "-")) %>%
  mutate(connections = {purrr::map(.$connections, ~combn(.x, 2))}) %>%
  rename(multiplet = multipletName) %>%
  addRowAndColumn(.) %>%
  addID(.) %>%
  mutate(expectedFrac = map(id, ~addExpectedFrac(.x)))

#filter plateData to only include those multiplets that passed initial QC
plateData <- plateData %>%
  filter(multiplet %in% rownames(getData(sObjPermutations, "spSwarm")))


known <- plateData %>%
  rename(multipletName = multiplet) %>%
  setupPlate(.)

#check the results, i.e. if the multiplet was correctly deconvoluted
res <- checkResults(sObjPermutations, known, 0) %>%
  select(multiplet, ACC) %>% 
  mutate(ACCbool = if_else(ACC == 1, TRUE, FALSE)) %>%
  select(-ACC)

#get permuted data
permData <- tidyPermutationData(sObjPermutations, permutations) 
```

The general goal here was to evaluate the "correctness" of the fractions that
are reported from the swarm deconvolution in comparison to different scenerios.
The sorted multiplets data, with an expected outcome, were used for the 
analysis.

We start by looking at the correspondance between the expected and 
observed fractions for each multiplet. For example, if we know that the 
multiplet is a doublet with HCT116 and HOS, we expect the fractions to be 0.5, 
0.5, 0, for HCT116, HOS, and A375, respectivley. Observed is, instead, the 
fractions reported by the swarm deconvolution.

```{r, fig.align='center', fig.height=8, fig.width=10, warning = FALSE}
#get the spSwarm fractions
dat <- sObjPermutations %>%
  getData(., "spSwarm") %>%
  rownames_to_column(var = "multiplet") %>%
  as_tibble() %>%
  #merge the plateData and move fractions into a list column
  inner_join(plateData, by = "multiplet") %>%
  full_join(res, by = "multiplet") %>%
  mutate(observedFrac = pmap(list(HOS, HCT116, A375), function(x, y, z) {
    c(x, y, z)
  })) %>%
  select(multiplet, multipletComposition, ACCbool, cellsInWell, expectedFrac, observedFrac) %>%
  unnest() %>%
  #calculate mean difference
  mutate(difference = observedFrac - expectedFrac) %>%
  arrange(cellsInWell, multipletComposition, difference) %>%
  mutate(cellsInWell = parse_factor(cellsInWell, levels = sort(unique(cellsInWell)))) %>%
  mutate(multiplet = parse_factor(multiplet, levels = unique(multiplet))) %>%
  mutate(multipletComposition = parse_factor(multipletComposition, levels = unique(multipletComposition)))

#plot
dat %>%
  ggplot() +
  geom_hline(yintercept = 0, lty = 2, colour = "lightgrey") +
  geom_hline(yintercept = .5, lty = 2, colour = "lightgrey") +
  geom_hline(yintercept = -.5, lty = 2, colour = "lightgrey") +
  geom_point(aes(
    x = multiplet, 
    y = difference, 
    fill = cellsInWell,
    colour = ACCbool
  ), size = 3, alpha = 0.75, shape = 21, lwd = 10) +
  facet_grid(. ~ multipletComposition, space = "free", scales = "free") +
  scale_fill_ptol() +
  scale_colour_manual(values = c("brown1", "black")) +
  scale_y_continuous(breaks = c(-0.5, 0, 0.5)) +
  theme_few() +
  labs(
    x = "Multiplet", 
    y = "Difference (observed fraction - expected fraction)"
  ) +
  theme(
    legend.position = "top", 
    plot.caption = element_text(hjust = 0),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    strip.text.x = element_text(angle = 90)
    ) +
  guides(
    colour = guide_legend(title = "Correctly deconvoluted"),
    fill = guide_legend(title = "# of cells in well")
  )
```

In the plot above we highlight multiplets which we know were deconvoluted 
incorrectly (via the oultine colour) since we expect these to have a larger
difference in fractions when compared to the expected fractions. 
Correct/incorrect deconvolution is judged by acccuracy = 1, or not. The plot 
confirms this in the majority of the cases. In A375-HOS-HOS multiplets we see
several cases where this is not the case. In one of these cases the observed 
fractions indicates that the majority of the multiplet is comprised of HOS. In 
the other case HCT116 has a non-zero fraction although it is quite small 
(0.00254).  

We can also look at the differences between the observed fractions and the 
fractions from permuted data.

```{r, fig.align='center', fig.height=8, fig.width=10}
permData %>%
  #add plate data
  full_join(select(plateData, multiplet, cellsInWell, multipletComposition), by = "multiplet") %>%
  #add accuracy measurment per multiplet
  full_join(res, by = "multiplet") %>%
  #add the cost p-value
  full_join(calculateCostP(permData), by = "multiplet") %>%
  mutate(pBool = if_else(pValue < 0.05, TRUE, FALSE)) %>%
  #expand the observed fractions to have the same dimensions as the permuted fractions
  #makes the difference calculation easier
  mutate(fracsExp = map(fracs, function(x) {
    as_tibble(x[rep(seq_len(nrow(x)), 10000), ])
  })) %>%
  #calculate difference between observed and permuted fractions
  mutate(differences = map2(fracsExp, permFracs, function(x, y) {
    y - x
  })) %>%
  #reformat to prepare for plotting
  select(multiplet, multipletComposition, ACCbool, cellsInWell, differences, pBool) %>%
  unnest() %>%
  gather(placeHolder, differences, -(multiplet:pBool)) %>%
  select(-placeHolder) %>%
  arrange(cellsInWell, multipletComposition, differences) %>%
  mutate(cellsInWell = parse_factor(cellsInWell, levels = sort(unique(cellsInWell)))) %>%
  mutate(multiplet = parse_factor(multiplet, levels = unique(multiplet))) %>%
  mutate(multipletComposition = parse_factor(multipletComposition, levels = unique(multipletComposition))) %>%
  #plot
  ggplot() +
  geom_hline(yintercept = 0, lty = 2, colour = "lightgrey") +
  geom_violin(aes(x = multiplet, y = differences, fill = cellsInWell, colour = ACCbool)) +
  facet_grid(. ~ multipletComposition, space = "free", scales = "free") +
  theme_few() +
  scale_fill_ptol() +
  scale_colour_manual(values = c("brown1", "black")) +
  scale_y_continuous(breaks = c(-1, 0, 1), labels = c("-1", "0", "1")) +
  labs(x = "Multiplet", y = "Difference (permuted fractions - observed fractions)") +
  guides(
    fill = guide_legend(title = "# of cells in well"),
    colour = guide_legend(title = "Correctly deconvoluted")
  ) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "top",
    strip.text.x = element_text(angle = 90)
  )
```


```{r}
sessionInfo()
```

