---
title: "Plot edges per multiplet"
output: html_document
---

The number of cells in a multiplet should correlate with the number of edges detected. Test this with the fetal pancreas dataset.
```{r}
library(sp.scRNAseq)
library(ggplot2)
library(ggthemes)
library(reshape2)

load('~/Desktop/sp.scRNAseqHalfTime/halfTime.rda')
load('~/Desktop/sp.scRNAseqHalfTime/sObj2.1000.rda')
load('~/Desktop/sp.scRNAseqHalfTime/sObj3.10000.rda')

#recalculate sObj so that costs are reported
sObj <- spSwarm(cObjMul, uObj, swarmsize=500, cores=4, max_iter=10, distFun=distToSlicePearson)

edges10 <- spSwarmPoisson(sObj, edge.cutoff = 1/13.5)
edges1000 <- spSwarmPoisson(sObj, edge.cutoff = 1/13.5)
edges10000 <- spSwarmPoisson(sObj, edge.cutoff = 1/13.5)

mulForEdges10 <- sapply(1:nrow(edges10), function(j)
    selectClustersOnEdge(sObj, 1/13.5, edges10[j, "from"], edges10[j, "to"])    
)

mulForEdges1000 <- sapply(1:nrow(edges1000), function(j)
    selectClustersOnEdge(sObj2, 1/13.5, edges1000[j, "from"], edges1000[j, "to"])    
)

mulForEdges10000 <- sapply(1:nrow(edges10000), function(j)
    selectClustersOnEdge(sObj3, 1/13.5, edges10000[j, "from"], edges10000[j, "to"])    
)

ta10 <- as.data.frame(table(unlist(mulForEdges10)))
ta1000 <- as.data.frame(table(unlist(mulForEdges1000)))
ta10000 <- as.data.frame(table(unlist(mulForEdges10000)))

#calculate cell number and frac.ercc
frac.ercc <- colSums(getData(cObjMul, "counts.ercc")) / (colSums(getData(cObjMul, "counts.ercc"))+colSums(getData(cObjMul, "counts")))

#calculate cell number
sfrac.ercc <- colSums(getData(cObjSng, "counts.ercc")) / (colSums(getData(cObjSng, "counts.ercc"))+colSums(getData(cObjSng, "counts")))
cellNr <- median(sfrac.ercc) / frac.ercc

if(identical(names(frac.ercc), names(cellNr))) {
    ercc <- data.frame(
        frac.ercc = frac.ercc,
        cellNr = cellNr
    )
}
ercc$Var1 <- rownames(ercc)
m <- Reduce(function(x, y) merge(x, y, all=TRUE, by="Var1", suffixes=c("10", "1000", "10000")), list(ta10, ta1000, ta10000, ercc))
colnames(m)[4] <- "Freq10000"
colnames(m)[1] <- "Sample"
mel <- reshape2::melt(m, id.vars=c("Sample", "cellNr", "frac.ercc"))

names <- c(
    Freq10 = "10 iterations",
    Freq1000 = "1000 iterations",
    Freq10000 = "10000 iterations"
)


ggplot(mel, aes(cellNr, value))+
    geom_point()+
    theme_few()+
    labs(
        x="Number of cells",
        y="Number of edges"
    )+
    facet_grid(variable~., labeller = as_labeller(names))

```

It looks as though the number of edges found is reduced between interation number 10 and 1000, although o further changes are seen when the iterations are increased to 10^4. Although at iteration 1000 and 10^4 there is no strong correlation between edges detected and number of cells, the somewhat negative correlation observed at iteration = 10, is absent. See if the sum of residuals for each multiplet differ between iterations.

```{r}
res10 <- calcResiduals(cObjMul, uObj, sObj)
res1000 <- calcResiduals(cObjMul, uObj, sObj2)
res10000 <- calcResiduals(cObjMul, uObj, sObj3)

cs <- data.frame(
    cs10=colSums(res10),
    cs1000=colSums(res1000),
    cs10000=colSums(res10000),
    mul=colnames(res10)
)

m2 <- reshape2::melt(cs, id.vars="mul")

ggplot(m2, aes(x=mul, y=value))+
    geom_bar(aes(fill=variable), stat="identity", position=position_dodge(width=1))+
    theme(
        axis.text.x=element_text(angle=90)
    )
```

The sum of residuals seems not to differ between iterations which is strange. Look at all the residuals individually.

```{r}
m10 <- melt(res10)
m1000 <- melt(res1000)
m10000 <- melt(res10000)
m10$variable <- 10
m1000$variable <- 1000
m10000$variable <- 10000

ms <- rbind(m10, m1000, m10000)

ggplot(ms, aes(Var2, value))+
    geom_point()+
    facet_grid(variable~.)+
    theme_few()+
    theme(
        axis.text.x=element_text(angle=90)
    )
```

This produced results I don't understand at the moment. Come back to this. 
Look at the costs reported instead for now.

```{r}
c10 <- getData(sObj, "costs")
c1000 <- getData(sObj2, "costs")
c10000 <- getData(sObj3, "costs")

d <- data.frame(
    c10=c10,
    c1000=c1000,
    c10000=c10000,
    mult=rownames(getData(sObj, "spSwarm"))
)

m3 <- melt(d, id.vars="mult")
ggplot(m3, aes(mult, value))+
    geom_bar(aes(fill=variable), stat="identity", position=position_dodge(width=1))+
    theme_few()+
    scale_fill_ptol()+
    theme(
        axis.text.x=element_text(angle=90)
    )
```

It appears that the costs are higher for iterations=10 than iterations=1000 or iterations=10000. Iterations 1000 and 10000 give the same costs as could potentially be expected.
Re-plot this as the sum of costs instead.

```{r}
sums <- as.data.frame(colSums(d[,1:3]))
sums$iterations <- rownames(sums)
colnames(sums)[1] <- "costs"

ggplot(sums, aes(x=iterations, y=costs))+
    geom_bar(stat="identity", position=position_dodge(width=1))+
    theme_few()+
    scale_x_discrete(labels=c("10", "1000", "10000"))+
    labs(
        y="Sum of costs"
    )


```