---
title: "Mouse analysis"
author: "Jason T. Serviss"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r, message = FALSE}
packages <- c("CIMseq", "sp.scRNAseqData", "tidyverse")
purrr::walk(packages, library, character.only = TRUE)
rm(packages)

##DATA
load('../data/output.rda')
load('../../MGA.analysis_SI/data/CIMseqData_181126.rda')
```

```{r}
#rename classes
renameClasses <- function(class) {
  case_when(
    class == "0" ~ "SI.TA.Mki67.medium",
    class == "1" ~ "SI.TA.Mki67.low",
    class == "2" ~ "SI.Stem",
    class == "3" ~ "SI.Goblet",
    class == "4" ~ "SI.TA.enterocyte",
    class == "5" ~ "SI.TA.Mki67.high.1",
    class == "6" ~ "SI.Enterocyte",
    class == "7" ~ "SI.TA.Mki67.high.2",
    class == "8" ~ "SI.Paneth",
    class == "9" ~ "SI.Tufft",
    class == "10" ~ "SI.Chromaffin",
    class == "11" ~ "Blood",
    TRUE ~ "error"
  )
}


getData(cObjSng.hq, "classification") <- renameClasses(getData(cObjSng.hq, "classification"))
getData(cObjSng, "classification") <- renameClasses(getData(cObjSng, "classification"))
fractions <- getData(sObj, "fractions")
colnames(fractions) <- renameClasses(colnames(fractions))
sObj@fractions <- fractions
```

```{r, fig.align="center", fig.width=10, fig.height=8}
cbind(getData(cObjSng.hq, "dim.red"), getData(cObjSng.hq, "classification")) %>%
  matrix_to_tibble("sample") %>%
  mutate(
    V1 = as.numeric(V1),
    V2 = as.numeric(V2)
  ) %>%
  ggplot() +
  geom_point(aes(V1, V2, colour = V3)) +
  scale_colour_manual(values = col40())
```

```{r}
spSwarmPoisson(sObj, cObjSng, 0.03)
```


Show table of number of cells detected per multiplet with edge cutoff = 0.01.
```{r}
cut1 <- apply(fractions, 1, function(r) {
  length(which(r > 0.01))
})
table(cut1)
```

Show table of number of cells detected per multiplet with edge cutoff = 0.001.
```{r}
cut2 <- apply(fractions, 1, function(r) {
  length(which(r > 0.001))
})
table(cut2)
```

Show histogram of fractions.
```{r, fig.align="center", fig.width=10, fig.height=8, eval = FALSE}
tibble(fractions = c(fractions)) %>%
  mutate(cut = if_else(fractions > 0.001, TRUE, FALSE)) %>%
  ggplot() +
  geom_histogram(aes(fractions), binwidth = 0.01) +
  facet_wrap(~cut, scales = "free")
```

Show cost per number of detected connections. Connections quantified with edge 
cutoff 0.01.
```{r}
cut1 <- apply(fractions, 1, function(r) {
  length(which(r > 0.01))
})

tibble(nConnections = cut1, cost = getData(sObj, "costs")) %>%
  mutate(nConnections = parse_factor(nConnections, levels = sort(unique(nConnections)))) %>%
  ggplot() + 
  geom_boxplot(aes(nConnections, cost))
```

Show ERCC fraction vs. detected connections

```{r}
ercc <- filter(estimateCells(cObjSng, cObjMul), sampleType == "Multiplet")
nConnections <- cut1[match(ercc$sampleName, names(cut1))]
tibble(
  cellNumber = ercc$cellNumberMedian,
  nConnections = nConnections
) %>%
  ggplot() +
  geom_boxplot(aes(nConnections, ercc, group = nConnections)) +
  labs(x = "n Connections", y = "ERCC fraction")
```

```{r, eval = FALSE}
ideal <- c(
  "SI.Goblet", "SI.Paneth", "SI.Stem", "SI.TA.Mki67.low", "SI.TA.Mki67.medium",
  "SI.TA.Mki67.high.1", "SI.TA.Mki67.high.2", "SI.TA.enterocyte", "SI.Enterocyte",
  "SI.Chromaffin", "SI.Tufft", "Blood"
)
cut <- 0.01
o <- tibble(sample = rownames(fractions)) %>%
  mutate(connections = map(sample, function(s) {
    sample <- fractions[s, ]
    colnames(fractions)[sample >= cut]
  })) %>%
  mutate(ctext = map_chr(connections, paste, collapse = ", ")) %>%
  mutate(cellNumber = map_int(connections, length)) %>%
  filter(cellNumber == 2) %>%
  mutate(ord = map_int(connections, function(c) {
    match(c[1], ideal)
  })) %>%
  mutate(ord2 = map_int(connections, function(c) {
    sum(match(c, ideal))
  })) %>%
  mutate(ord3 = map2_chr(sample, connections, function(s, c) {
    fractions[s, c[1]]
  })) %>%
  arrange(ord, cellNumber, ord2, ord3) 

fractions[pull(o, sample), ] %>%
  matrix_to_tibble("sample") %>%
  mutate(sample = parse_factor(sample, levels = unique(sample))) %>%
  gather(class, frac, -sample) %>%
  mutate(class = parse_factor(class, levels = ideal)) %>%
  group_by(sample) %>%
  mutate(cn = length(which(frac > cut))) %>%
  ungroup() %>%
  ggplot() +
  geom_tile(aes(class, sample, fill = frac)) +
  #facet_wrap( ~ cn, scales = "free_x") +
  scale_fill_viridis_c() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  ) +
  coord_flip()
```

```{r, eval = FALSE}
#a quick test with 2 samples to confirm plotting is working as expected
ideal <- c(
  "SI.Goblet", "SI.Paneth", "SI.Stem", "SI.TA.Mki67-low", "SI.TA.Mki67-medium",
  "SI.TA.Mki67-high.1", "SI.TA.Mki67-high.2", "SI.TA.enterocyte", "SI.Enterocyte",
  "SI.Chromaffin", "SI.Tufft", "Blood"
)
fractions <- getData(sObj, "fractions")
test <- c("m.NJA01301.F22", "m.NJA01202.J20")

cut <- 0.01
tibble(sample = test) %>%
  mutate(from = map(sample, function(s) {
    sample <- fractions[s, ]
    n <- colnames(fractions)[sample > cut]
    if(length(n) == 1) {
      tibble(Var1 = n, Var2 = n)
    } else {
      expand.grid(n, n, stringsAsFactors = FALSE) %>%
        as.tibble()
    }
  })) %>%
  unnest() %>%
  filter(Var1 != Var2) %>%
  inner_join(matrix_to_tibble(fractions, "sample"), by = "sample") %>%
  gather(class, frac, -(sample:Var2)) %>%
  mutate(frac = pmap_dbl(list(Var1, Var2, class, frac), function(fr, to, c, f) {
    if_else(c %in% c(fr, to), f, 0)
  })) %>% 
  mutate(
    xord1 = match(Var1, ideal),
    xord2 = match(Var2, ideal)
  ) %>%
  arrange(xord1, xord2, frac) %>%
  mutate(x = paste(sample, Var1, Var2, sep = "-")) %>%
  mutate(x = parse_factor(x, levels = unique(x))) %>%
  mutate(class = parse_factor(class, levels = ideal)) %>%
  mutate(Var1 = parse_factor(Var1, levels = ideal)) %>%
  inner_join(tibble(
    sample = rownames(getData(sObj, "fractions")), 
    detectedCon = apply(getData(sObj, "fractions"), 1, function(r) {
      length(which(r > 0.01))
    })
  )) %>%
  ggplot() +
  geom_tile(aes(x, class, fill = frac)) +
  facet_wrap(~ Var1, scales = "free_x") +
  labs(x = "Connections", y = "Class") +
  scale_fill_viridis_c() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  ) +
  guides(fill = guide_colorbar(title = "Fractions"))
```

Plot results for all multiplets.

```{r, fig.align="center", fig.width=12, fig.height=10}
ideal <- c(
  "SI.Goblet", "SI.Paneth", "SI.Stem", "SI.TA.Mki67.low", "SI.TA.Mki67.medium",
  "SI.TA.Mki67.high.1", "SI.TA.Mki67.high.2", "SI.TA.enterocyte", "SI.Enterocyte",
  "SI.Chromaffin", "SI.Tufft", "Blood"
)

cut <- 0.1
tibble(sample = rownames(fractions)) %>%
  mutate(from = map(sample, function(s) {
    sample <- fractions[s, ]
    n <- colnames(fractions)[sample > cut]
    if(length(n) == 1) {
      tibble(Var1 = n, Var2 = n)
    } else {
      expand.grid(n, n, stringsAsFactors = FALSE) %>%
        as.tibble()
    }
  })) %>%
  unnest() %>%
  filter(Var1 != Var2) %>%
  inner_join(matrix_to_tibble(fractions, "sample"), by = "sample") %>%
  gather(class, frac, -(sample:Var2)) %>%
  mutate(frac = pmap_dbl(list(Var1, Var2, class, frac), function(fr, to, c, f) {
    if_else(c %in% c(fr, to), f, 0)
  })) %>% 
  mutate(
    xord1 = match(Var1, ideal),
    xord2 = match(Var2, ideal)
  ) %>%
  arrange(xord1, xord2, frac) %>%
  mutate(x = paste(sample, Var1, Var2, sep = "-")) %>%
  mutate(x = parse_factor(x, levels = unique(x))) %>%
  mutate(class = parse_factor(class, levels = ideal)) %>%
  mutate(Var1 = parse_factor(Var1, levels = ideal)) %>%
  ggplot() +
  geom_tile(aes(x, class, fill = frac)) +
  facet_wrap(~ Var1, scales = "free_x") +
  labs(x = "Connections", y = "Class") +
  scale_fill_viridis_c() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  ) +
  guides(fill = guide_colorbar(title = "Fractions"))
```

Plot results for multiplets with duplicates detected.

```{r, fig.align="center", fig.width=12, fig.height=10}
ideal <- c(
  "SI.Goblet", "SI.Paneth", "SI.Stem", "SI.TA.Mki67.low", "SI.TA.Mki67.medium",
  "SI.TA.Mki67.high.1", "SI.TA.Mki67.high.2", "SI.TA.enterocyte", "SI.Enterocyte",
  "SI.Chromaffin", "SI.Tufft", "Blood"
)

cut <- 0.01
tibble(sample = rownames(fractions)) %>%
  mutate(from = map(sample, function(s) {
    sample <- fractions[s, ]
    n <- colnames(fractions)[sample >= cut]
    if(length(n) == 1) {
      tibble(Var1 = n, Var2 = n)
    } else {
      expand.grid(n, n, stringsAsFactors = FALSE) %>%
        as.tibble()
    }
  })) %>%
  unnest() %>%
  full_join(matrix_to_tibble(fractions, "sample"), by = "sample") %>%
  gather(class, frac, -(sample:Var2)) %>%
  mutate(frac = pmap_dbl(list(Var1, Var2, class, frac), function(fr, to, c, f) {
    if_else(c %in% c(fr, to), f, 0)
  })) %>% 
  mutate(
    xord1 = match(Var1, ideal),
    xord2 = match(Var2, ideal)
  ) %>%
  arrange(xord1, xord2, frac) %>%
  mutate(x = paste(sample, Var1, Var2, sep = "-")) %>%
  mutate(x = parse_factor(x, levels = unique(x))) %>%
  mutate(class = parse_factor(class, levels = ideal)) %>%
  mutate(Var1 = parse_factor(Var1, levels = ideal)) %>%
  full_join(tibble(
    sample = rownames(getData(sObj, "fractions")), 
    detectedCon = apply(getData(sObj, "fractions"), 1, function(r) {
      length(which(r > 0.01))
    })
  )) %>%
  filter(detectedCon == 2) %>%
  ggplot() +
  geom_tile(aes(x, class, fill = frac)) +
  facet_wrap(~ Var1, scales = "free_x") +
  labs(x = "Connections", y = "Class") +
  scale_fill_viridis_c() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  ) +
  guides(fill = guide_colorbar(title = "Fractions"))
```

Plot results for multiplets with triplicates detected.

```{r, fig.align="center", fig.width=12, fig.height=10}
ideal <- c(
  "C.Goblet.distal", "C.Stem.distal", "C.TA.distal", "C.Colonocyte",
  "C.TA.proximal.1", "C.TA.proximal.2", "C.Stem.proximal", "C.Goblet.proximal",
  "Chromaffin", "Tufft", "Blood", "SI.Goblet", "SI.Paneth", "SI.Stem.1",
  "SI.Stem.2", "SI.TA.early.1", "SI.TA.early.2", "SI.TA.intermediate",
  "SI.TA.late", "SI.Enterocyte"
)

cut <- 0.001
tibble(sample = rownames(fractions)) %>%
  mutate(from = map(sample, function(s) {
    sample <- fractions[s, ]
    n <- colnames(fractions)[sample >= cut]
    if(length(n) == 1) {
      tibble(Var1 = n, Var2 = n)
    } else {
      expand.grid(n, n, stringsAsFactors = FALSE) %>%
        as.tibble()
    }
  })) %>%
  unnest() %>%
  full_join(matrix_to_tibble(fractions, "sample"), by = "sample") %>%
  gather(class, frac, -(sample:Var2)) %>%
  mutate(frac = pmap_dbl(list(Var1, Var2, class, frac), function(fr, to, c, f) {
    if_else(c %in% c(fr, to), f, 0)
  })) %>% 
  mutate(
    xord1 = match(Var1, ideal),
    xord2 = match(Var2, ideal)
  ) %>%
  arrange(xord1, xord2, frac) %>%
  mutate(x = paste(sample, Var1, Var2, sep = "-")) %>%
  mutate(x = parse_factor(x, levels = unique(x))) %>%
  mutate(class = parse_factor(class, levels = ideal)) %>%
  mutate(Var1 = parse_factor(Var1, levels = ideal)) %>%
  full_join(tibble(
    sample = rownames(getData(sObj, "fractions")), 
    detectedCon = apply(getData(sObj, "fractions"), 1, function(r) {
      length(which(r > 0.01))
    })
  )) %>%
  filter(detectedCon == 3) %>%
  ggplot() +
  geom_tile(aes(x, class, fill = frac)) +
  facet_wrap(~ Var1, scales = "free_x") +
  labs(x = "Connections", y = "Class") +
  scale_fill_viridis_c() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  ) +
  guides(fill = guide_colorbar(title = "Fractions"))
```

```{r, fig.align="center", fig.width=12, fig.height=10, eval = FALSE}


edgeDat <- getEdgesForMultiplet(sObj, 0.001, rownames(fractions)) %>%
  group_by(multiplet) %>%
  summarize(edges = paste(paste(from, to, sep = "-"), collapse = ", "))

p <- fractions %>%
  matrix_to_tibble("sample") %>%
  gather(class, fraction, -sample) %>%
  mutate(classTissue = case_when(
    str_detect(class, "^C\\.") ~ "colon", 
    str_detect(class, "^S") ~ "SI",
    str_detect(class, "^[T, C, B]") ~ "miscellaneous", 
    TRUE ~ "error"
  )) %>%
  inner_join(MGA.Meta, by = "sample") %>%
  inner_join(edgeDat, by = c("sample" = "multiplet")) %>%
  arrange(edges) %>%
  mutate(sample = parse_factor(sample, levels = unique(sample))) %>%
  ggplot() +
  geom_tile(aes(class, sample, fill = log2(fraction + 1))) +
  facet_grid(. ~ classTissue, scales = "free") +
  scale_fill_viridis_c() +
  theme_bw() +
  labs(x = "Samples", caption = "edge cutoff = 0") +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.title.x = element_blank(),
    axis.text.x = element_text(angle = 90),
    legend.position = "top",
    plot.caption = element_text(hjust = 0)
  ) +
  guides(fill = guide_colourbar(title = "Fraction", title.position = "top", title.hjust = 0.5))

p
```

```{r, fig.align="center", fig.width=12, fig.height=10, eval = FALSE}
p <- round(swarm, digits = 2) %>%
  add_column(cd = apply(., 1, function(f) length(which(f != 0)))) %>%
  rownames_to_column("sample") %>%
  as_tibble() %>%
  gather(class, fraction, -sample, -cd) %>%
  mutate(classTissue = case_when(
    class %in% c("5", "1", "15", "6", "4") ~ "colon", 
    class %in% c("10", "13", "14", "2", "3", "7", "8", "9") ~ "SI",
    class %in% c("12", "11") ~ "miscellaneous", 
    TRUE ~ "error"
  )) %>%
  inner_join(MGA.Meta, by = "sample") %>%
  mutate(class = renameClasses(class)) %>%
  inner_join(edgeDat, by = c("sample" = "multiplet")) %>%
  arrange(edges) %>%
  mutate(sample = parse_factor(sample, levels = unique(sample))) %>%
  ggplot() +
  geom_tile(aes(class, sample, fill = log2(fraction + 1))) +
  facet_grid(cd ~ classTissue, scales = "free") +
  scale_fill_viridis_c() +
  theme_bw() +
  labs(x = "Samples", caption = "edge cutoff = 0.01") +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.title.x = element_blank(),
    axis.text.x = element_text(angle = 90),
    legend.position = "top",
    plot.caption = element_text(hjust = 0)
  ) +
  guides(fill = guide_colourbar(title = "Fraction", title.position = "top", title.hjust = 0.5))

p
```

```{r, fig.align="center", fig.width=12, fig.height=10, eval = FALSE}
spSwarmPoisson(sObj, 0.001) %>%
  filter(from != "undefined" & to != "undefined") %>%
  arrange(from, to) %>%
  mutate(
    from.name = renameClasses(from),
    to.name = renameClasses(to)
  ) %>%
  mutate(
    from.name = parse_factor(from.name, levels = unique(from.name)),
    to.name = parse_factor(to.name, levels = unique(to.name))
  ) %>%
  ggplot() +
  geom_tile(aes(from.name, to.name, fill = weight)) +
  scale_fill_viridis_c() +
  labs(x = "from", y = "to", caption = "edge.cutoff = 0.01") +
  theme(
    plot.caption = element_text(hjust = 0),
    axis.text.x = element_text(angle = 90),
    legend.position = "top"
  ) +
  guides(fill = guide_colourbar(title = "Weight", title.position = "top", title.hjust = 0.5))
```

```{r, fig.align="center", fig.width=12, fig.height=10, eval = FALSE}
weights <- spSwarmPoisson(sObj, 0.1)

freq <- expand.grid(colnames(fractions), colnames(fractions), stringsAsFactors = FALSE) %>%
  as_tibble() %>%
  mutate(weight = map2_int(Var1, Var2, function(f, t) {
    filter(weights, (from == f & to == t) | (from == t & to == f))$weight
  })) %>%
  group_by(Var1) %>%
  mutate(totalWeights = sum(weight)) %>%
  ungroup() %>%
  mutate(Frequency = weight / totalWeights)

data <- spSwarmPoisson(sObj, 0.1) %>%
  filter(from != "undefined" & to != "undefined") %>%
  arrange(from, to) %>%
  mutate(
    from.name = parse_factor(from, levels = unique(from)),
    to.name = parse_factor(to, levels = unique(to))
  ) %>%
  inner_join(freq, by = c("from" = "Var1", "to" = "Var2"))


data %>%
  ggplot() +
  geom_tile(aes(from.name, to.name, fill = Frequency)) +
  scale_fill_viridis_c() +
  labs(x = "from", y = "to", caption = "edge.cutoff = 0.01; 'Frequency' is the frequency of 'from' interaction with 'to'.") +
  theme(
    plot.caption = element_text(hjust = 0),
    axis.text.x = element_text(angle = 90),
    legend.position = "top"
  ) +
  guides(fill = guide_colourbar(title = "Frequency", title.position = "top", title.hjust = 0.5))
```


####NOV 12
```{r}
library(ggalluvial)
```

```{r}
tibble(sample = rownames(fractions)) %>%
  mutate(connections = map(sample, function(s) {
    sample <- fractions[s, ]
    c <- colnames(fractions)[sample >= cut]
    if(length(c) == 1) {
      tibble(from = c, to = c)
    } else {
      combn(c, 2) %>%
        t() %>%
        as_tibble() %>%
        rename(from = V1, to = V2)
    }
  })) %>%
  mutate(nConnections = map_dbl(connections, nrow)) %>%
  unnest() %>%
  mutate(sum = pmap_dbl(list(sample, from, to), function(s, f, t) {
    sum(fractions[s, c(f, t)])
  })) %>%
  mutate(
    from = parse_factor(from, levels = sort(unique(from))),
    to = parse_factor(to, levels = sort(unique(to)))
  ) %>%
  ggplot(aes(axis1 = from, axis2 = to)) +
  geom_alluvium(aes(fill = sum)) + 
  geom_stratum(width = 0.1) + 
  geom_label(stat = "stratum", label.strata = TRUE) +
  scale_fill_viridis() +
  ggthemes::theme_few()
```

```{r}
calculateBinomP <- function(swarm, edge.cutoff) {
  fractions <- getData(swarm, "fractions")
  totals <- apply(fractions, 2, function(f) length(which(f > edge.cutoff)))
  classes <- colnames(fractions)
  observed <- spSwarmPoisson(sObj, edge.cutoff)

  expand.grid(classes, classes, stringsAsFactors = FALSE) %>%
    rename(from = Var1, to = Var2) %>%
    mutate(weight = map2_dbl(from, to, function(f, t) {
      if(f == t) {
        filter(observed, from == f & to == t)$weight
      } else {
        bool <- (observed$from == f | observed$to == f) & (observed$from == t | observed$to == t)
        observed[bool, "weight"][[1]]
      }
    })) %>%
    mutate(total.from = map_dbl(from, ~totals[.x])) %>%
    mutate(expected.from = total.from / length(classes)) %>%
    mutate(prob = 1/ total.from) %>%
    mutate(score = weight / expected.from) %>%
    mutate(p.value = pmap_dbl(list(weight, total.from, prob), function(o, t, p) {
      binom.test(o, t, p = p, alternative = "greater")$p.value
    }))
}
```

```{r}
calculateBinomP(sObj, 0.01) %>%
  as_tibble() %>%
  filter(p.value < 0.01) %>%
  filter(weight > 30) %>%
  ggplot(aes(y = score, axis1 = from, axis2 = to)) +
  geom_alluvium(aes(fill = from)) + 
  geom_stratum(width = 0.1) + 
  geom_label(stat = "stratum", label.strata = TRUE) +
  #scale_fill_viridis() +
  scale_fill_manual(values = col40()) +
  ggthemes::theme_few() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank()
  ) + 
  guides(fill = FALSE)
```

#Nov 13
```{r}
library(circlize)
```

```{r}
ideal <- c(
  "SI.Goblet", "SI.Paneth", "SI.Stem", "SI.TA.Mki67.low", "SI.TA.Mki67.medium",
  "SI.TA.Mki67.high.1", "SI.TA.Mki67.high.2", "SI.TA.enterocyte", "SI.Enterocyte",
  "SI.Chromaffin", "SI.Tufft", "Blood"
)

cut <- 0.01
data <- tibble(sample = rownames(fractions)) %>%
  mutate(from = map(sample, function(s) {
    sample <- fractions[s, ]
    n <- colnames(fractions)[sample > cut]
    if(length(n) == 1) {
      tibble(from = n, to = n, nConnections = 0)
    } else {
      combn(n, 2) %>%
        t() %>%
        as.tibble() %>%
        rename(from = V1, to = V2) %>%
        mutate(nConnections = ncol(combn(n, 2)))
    }
  })) %>%
  unnest() %>%
  filter(from != to) %>% #removes self connections
  mutate(frac.1 = map2_dbl(sample, from, function(s, c) {
    fractions[s, c]
  })) %>%
  mutate(frac.2 = map2_dbl(sample, to, function(s, c) {
    fractions[s, c]
  })) %>%
  mutate(
    ord1 = match(from, ideal),
    ord2 = match(to, ideal)
  ) %>%
  arrange(ord1, ord2, frac.1) %>%
  group_by(from) %>%
  mutate(from.n = normalizeVec(1:n())) %>%
  ungroup() %>%
  group_by(to) %>%
  mutate(to.n = normalizeVec(1:n())) %>%
  ungroup() %>%
  as.data.frame()

#plot

widths <- table(c(data$from, data$to))[ideal]
widths["SI.Tufft"] <- widths["SI.Tufft"] + 50
widths["SI.Chromaffin"] <- widths["SI.Chromaffin"] + 75
widths["SI.Enterocyte"] <- widths["SI.Enterocyte"] + 75
widths["Blood"] <- widths["Blood"] + 25
widths["SI.Paneth"] <- widths["SI.Paneth"] + 25
circos.initialize(
  factors = ideal, 
  xlim = c(0, 1), sector.width = widths
)

circos.trackPlotRegion(
  ylim = c(0, 1), bg.col = col40()[1:length(ideal)], 
  bg.border = NA, track.height = 0.1
)

#add labels
for(n in ideal) {
  circos.text(
    0.5, 0.5, n, n, 1, col = "white", facing = "bending.inside", cex = 0.8
  )
}

#add fraction heatmap
cols <-  viridis::viridis(500)
heat <- tibble(
  class = c(data$from, data$to),
  frac = c(data$frac.1, data$frac.2),
  position = c(data$from.n, data$to.n)
) %>%
  distinct() %>%
  mutate(idx = ntile(frac, nrow(.))) %>%
  mutate(color = viridis::viridis(max(idx))[idx]) %>%
  arrange(class, position)

circos.track(
  ylim = c(0, 1), bg.border = "darkgrey", track.height = 0.05, track.margin = c(0.0001, 0.0001),
  panel.fun = function(x, y) {
    sector.index = CELL_META$sector.index
    m <- filter(heat, class == sector.index)
    
    nr <- nrow(m)
    nc <- ncol(m)
    for(i in 1:nr) {
      circos.rect(
        xleft = m$position[i], ybottom = 0, 
        xright = m$position[i], ytop = 1, 
        sector.index = sector.index,
        border = m$color[i], col = m$color[i]
      )
    }
})

#add links
for(i in 1:nrow(data)) {
  circos.link(
    data[i, "from"], data[i, "from.n"],
    data[i, "to"], data[i, "to.n"],
    border = 1, col = rgb(0, 0, 0, 0.2)
  )
}



```


```{r}
ideal <- c(
  "SI.Goblet", "SI.Paneth", "SI.Stem", "SI.TA.Mki67.low", "SI.TA.Mki67.medium",
  "SI.TA.Mki67.high.1", "SI.TA.Mki67.high.2", "SI.TA.enterocyte", "SI.Enterocyte",
  "SI.Chromaffin", "SI.Tufft", "Blood"
)

cut <- 0.01
edges <- getEdgesForMultiplet(sObj, cut, rownames(fractions))
data <- edges %>%
  filter(from == "SI.Paneth" | to == "SI.Paneth") %>%
  rename(sample = multiplet) %>%
  filter(from != to) %>% #removes self connections
  group_by(from, to) %>%
  filter(n() > 5) %>%
  ungroup() %>%
  mutate(
    frac.1 = map2_dbl(sample, from, function(s, c) {
      fractions[s, c]
    }),
    frac.2 = map2_dbl(sample, to, function(s, c) {
      fractions[s, c]
    })
  ) %>%
  mutate(
    ord1 = match(from, ideal),
    ord2 = match(to, ideal)
  ) %>%
  arrange(ord1, ord2, frac.1) %>%
  select(-ord1, -ord2) %>%
  as.data.frame() %>%
  mutate(connectionID = 1:nrow(.)) %>%
  mutate(connectionType = paste(from, to, sep = "--"))

heat <- tibble(
  class = c(data$from, data$to),
  frac = c(data$frac.1, data$frac.2),
  connectionID = c(data$connectionID, data$connectionID),
  connectionType = c(data$connectionType, data$connectionType),
  sample = c(data$sample, data$sample)
) %>%
  arrange(class, connectionType) %>%
  group_by(class) %>%
  mutate(position = 1:n())

#plot
class.colors <- col40()[1:length(ideal)]
names(class.colors) <- ideal

widths <- table(heat$class)[unique(heat$class)]
widths["Blood"] <- widths["Blood"] + 2

widths["SI.Tufft"] <- widths["SI.Tufft"] + 55
widths["SI.Chromaffin"] <- widths["SI.Chromaffin"] + 80
widths["SI.Enterocyte"] <- widths["SI.Enterocyte"] + 50

widths["SI.Paneth"] <- widths["SI.Paneth"] + 25

circos.par(cell.padding = c(0.02, 0, 0.02, 0))
circos.initialize(factors = heat$class, x = heat$position, sector.width = widths[sort(unique(heat$class))])
circos.trackPlotRegion(
  ylim = c(0, 1), bg.col = class.colors[sort(unique(heat$class))], 
  bg.border = NA, track.height = 0.1
)

#add labels
for(n in unique(heat$class)) {
  x <- mean(range(filter(heat, class == n)$position))
  circos.text(x, 0.5, n, n, 1, col = "white", facing = "bending.inside", cex = 0.8)
}

#add links
alpha.colors <- ggplot2::alpha(class.colors, 0.2)
names(alpha.colors) <- names(class.colors)
for(i in 1:length(unique(heat$connectionID))) {
  conn <- filter(heat, connectionID == sort(unique(heat$connectionID))[i])
  if(nrow(conn) != 2) stop("error")
  circos.link(
    pull(conn, class)[1], pull(conn, position)[1],
    pull(conn, class)[2], pull(conn, position)[2],
    border = 1, col = rgb(0, 0, 0, 0.2)
  )
}
```

```{r}
ideal <- c(
  "SI.Goblet", "SI.Paneth", "SI.Stem", "SI.TA.Mki67.low", "SI.TA.Mki67.medium",
  "SI.TA.Mki67.high.1", "SI.TA.Mki67.high.2", "SI.TA.enterocyte", "SI.Enterocyte",
  "SI.Chromaffin", "SI.Tufft", "Blood"
)

cut <- 0.01
edges <- getEdgesForMultiplet(sObj, cut, rownames(fractions))
data <- edges %>%
  filter(from == "SI.Paneth" | to == "SI.Paneth") %>%
  rename(sample = multiplet) %>%
  filter(from != to) %>% #removes self connections
  group_by(from, to) %>%
  filter(n() > 5) %>%
  ungroup() %>%
  mutate(
    frac.1 = map2_dbl(sample, from, function(s, c) {
      fractions[s, c]
    }),
    frac.2 = map2_dbl(sample, to, function(s, c) {
      fractions[s, c]
    })
  ) %>%
  mutate(
    ord1 = match(from, ideal),
    ord2 = match(to, ideal)
  ) %>%
  arrange(ord1, ord2, frac.1) %>%
  select(-ord1, -ord2) %>%
  as.data.frame() %>%
  mutate(connectionID = 1:nrow(.)) %>%
  mutate(connectionType = map2_chr(from, to, ~paste(sort(c(.x, .y)), collapse = "--")))

heat <- tibble(
  class = c(data$from, data$to),
  frac = c(data$frac.1, data$frac.2),
  connectionID = c(data$connectionID, data$connectionID),
  connectionType = c(data$connectionType, data$connectionType),
  sample = c(data$sample, data$sample)
) %>%
  arrange(class, connectionType, frac) %>%
  group_by(class) %>%
  mutate(position = 1:n()) %>%
  ungroup() %>%
  mutate(idx = ntile(frac, nrow(.))) %>%
  mutate(color = viridis::viridis(max(idx))[idx])

#plot
class.colors <- col40()[1:length(ideal)]
names(class.colors) <- ideal

circos.par(cell.padding = c(0.02, 0, 0.02, 0))
circos.initialize(factors = unique(c(edges$from, edges$to)), xlim = range(heat$position))
circos.trackPlotRegion(
  ylim = c(0, 1), bg.col = class.colors[sort(unique(c(edges$from, edges$to)))], 
  bg.border = NA, track.height = 0.1
)

#add labels
for(n in unique(c(edges$from, edges$to))) {
  circos.text(
    mean(range(heat$position)), 0.5, n, n, 1, col = "white", 
    facing = "bending.inside", cex = 0.8
  )
}

#add fractions
circos.track(
  ylim = c(0, 1), bg.border = "darkgrey", track.height = 0.05, track.margin = c(0.0001, 0.0001),
  panel.fun = function(x, y) {
    sector.index = CELL_META$sector.index
    m <- filter(heat, class == sector.index)
    if(nrow(m) == 0) return(NA)
    for(i in 1:nrow(m)) {
      circos.rect(
        xleft = m$position[i], ybottom = 0, 
        xright = m$position[i], ytop = 1, 
        sector.index = sector.index,
        border = m$color[i], col = m$color[i]
      )
    }
})

#add links
ps <- calculateBinomP(sObj, 0.01) %>%
  mutate(conn = map2_chr(from, to, ~paste(sort(c(.x, .y)), collapse = "--"))) %>%
  filter(conn %in% heat$connectionType) %>%
  mutate(idx = ntile(p.value, nrow(.))) %>%
  mutate(col = viridis::cividis(max(idx))[rev(idx)]) %>%
  arrange(idx)

for(i in 1:length(unique(heat$connectionID))) {
  conn <- filter(heat, connectionID == sort(unique(heat$connectionID))[i])
  c <- ps[ps$from == pull(conn, class)[1] & ps$to == pull(conn, class)[2], "col"]
  if(nrow(conn) != 2) stop("error")
  circos.link(
    pull(conn, class)[1], pull(conn, position)[1],
    pull(conn, class)[2], pull(conn, position)[2],
    border = 1, col = c
  )
}
```


```{r}
#TEST
ideal <- LETTERS[1:4]
fracs <- matrix(runif(12, 1, 10), ncol = 2)

data <- combn(ideal, 2) %>%
  t() %>%
  as.data.frame(stringsAsFactors = FALSE) %>%
  rename(from = V1, to = V2) %>%
  bind_rows(., .) %>%
  mutate(
    frac.1 = seq(1, nrow(.), 1),
    frac.2 = seq(nrow(.), 1, -1)
  ) %>%
  mutate(
    ord1 = match(from, ideal),
    ord2 = match(to, ideal)
  ) %>%
  arrange(ord1, ord2, frac.1) %>%
  mutate(from.n = seq(1, nrow(.), 1), to.n = seq(nrow(.), 1, -1))

heat <- tibble(
  class = c(data$from, data$to),
  frac = c(data$frac.1, data$frac.2),
  position = c(data$from.n, data$to.n)
) %>%
  group_by(class) %>%
  arrange(frac) %>%
  ungroup() %>%
  mutate(idx = 1:n()) %>%
  mutate(color = viridis::viridis(nrow(.))[idx])

#plot
circos.par(cell.padding = c(0.02, 0, 0.02, 0))
circos.initialize(factors = heat$class, x = heat$position)
circos.trackPlotRegion(
  ylim = c(0, 1), bg.col = col40()[1:length(ideal)], 
  bg.border = NA, track.height = 0.1
)

#add labels
for(n in ideal) {
  x <- mean(filter(heat, class == n)$position)
  circos.text(x, 0.5, n, n, 1, col = "white", facing = "bending.inside", cex = 0.8)
}

#add fraction heatmap

circos.track(
  ylim = c(0, 1), bg.border = "darkgrey", track.height = 0.05, panel.fun = function(x, y) {
    sector.index = CELL_META$sector.index
    m <- filter(heat, class == sector.index)
    
    nr <- nrow(m)
    nc <- ncol(m)
    for(i in 1:nr) {
      circos.rect(
        xleft = m$position[i] - 0.001, ybottom = 0, 
        xright = m$position[i] + 0.001, ytop = 1, 
        sector.index = sector.index,
        border = m$color[i], col = m$color[i]
      )
    }
})

#add links
for(i in 1:nrow(data)) {
  circos.link(
    data[i, "from"], data[i, "from.n"],
    data[i, "to"], data[i, "to.n"],
    border = 1, col = rgb(0, 0, 0, 0.2)
  )
}
```


#GRAPH

```{r}
p <- spSwarmPoisson(sObj, edge.cutoff = 0.01) %>%
    unite('connection', .data$from, .data$to, sep = "-", remove = FALSE) %>%
    select(.data$from, .data$to, .data$connection, .data$weight, .data$pval) %>%
    igraph::graph_from_data_frame(directed = FALSE) %>%
    tidygraph::as_tbl_graph() %>% #better if this could be done directly; avoids importing igraph
    tidygraph::activate(nodes) %>%
    rename('Class' = .data$name) %>%
    #remove edges with 0 weight and coerce to factor
    tidygraph::activate(edges) %>%
    filter(.data$weight > 0)

#layouts: drl, kk, grid, fr, dh, gem, lgl, mds, circle, randomly, graphopt, nicely
ps <- calculateBinomP(sObj, 0.01) %>%
  unite(connection.1, from, to, sep = "-", remove = FALSE) %>%
  unite(connection.2, to, from, sep = "-", remove = FALSE) %>%
  filter(p.value < 0.05)

p %>%
  tidygraph::activate(edges) %>%
  filter(connection %in% ps$connection.1 | connection %in% ps$connection.2) %>%
  ggraph(layout = 'linear', circular = TRUE) +
  #plot edges
  geom_edge_arc(
      edge_colour = "black", aes_string(edge_width = 'weight'), 
      edge_alpha = 0.3, lineend = "round"
  ) +
  # add all cells
  geom_node_point(aes(colour = Class), size = 4) +
  scale_colour_manual(name = "classification", values = col40()) +
  theme(legend.position = "top", legend.title.align = 0.5) +
  guides(
    colour = guide_legend(title = "Classification", title.position = "top"),
    edge_width = guide_legend(title = "Weight", title.position = "top")
  )
```

```{r}
data <- getEdgesForMultiplet(sObj, cut, rownames(fractions)) %>%
  mutate(connectionID = 1:n()) %>%
  select(-multiplet) %>%
  group_by(from, to) %>%
  mutate(weight = n()) %>%
  filter(weight > 5) %>%
  igraph::graph_from_data_frame(directed = FALSE) %>%
  tidygraph::as_tbl_graph()

set.seed(98342)
#layouts: drl, kk, grid, fr, dh, gem, lgl, mds, circle, randomly, graphopt, nicely
final_layout <- create_layout(graph = data, layout = "circle")

ggraph(data, layout = "manual", node.positions = final_layout) +
  geom_edge_fan(
    edge_colour = "black", edge_alpha = 0.2, lineend = "round"
  ) +
  geom_node_point(size = 4, aes(colour = name)) +
  scale_colour_manual(values = col40()) +
  guides(
    colour = guide_legend(title = "Classification", title.position = "top")
  )

```

